---
title: "HIV RNASeq"
author: "Guannan Shen"
date: "November 21, 2018"
output: 
  pdf_document:
    latex_engine: lualatex
    number_sections: yes
    toc: yes
    toc_depth: 5
  word_document:
    toc: yes
    toc_depth: '5'
  html_document:
    number_sections: yes
    theme: united
    toc: yes
    toc_depth: 5
    toc_float: yes
---

```{r setup, include=FALSE, cache = FALSE}
require("knitr")
opts_chunk$set(tidy.opts=list(width.cutoff=60),tidy=TRUE)
opts_chunk$set(engine = "R")
knitr::opts_chunk$set(echo = T)
knitr::opts_chunk$set(message = F)
knitr::opts_chunk$set(warning = F)
## setting wd in DELL
## opts_knit$set(root.dir = "~/Stats/CIDA_OMICs/CIDA_OMICS/7659Stats_Genetics/HW5/")
## setting working directory in asus 
## opts_knit$set(root.dir = "C:/Users/hithr/Documents/Stats/gitlab/Cario_RNASeq_Microbiom_Inte/DataRaw/") 
## setting working directory in ubuntu
opts_knit$set(root.dir = "~/Documents/gitlab/Cario_RNASeq_Microbiom_Inte/DataRaw/")
                                                 
## cache = F, if cache = T, will not revaluate code chunk everytime
## double or more space to insert a line break
```

```{r libs}
######################
## Set up workspace
######################
rm(list = ls())
library(edgeR)
library(EDASeq)
library(DESeq2)
library(knitr)
library(tidyverse)
library(magrittr)
library(stats)
library(BiocParallel)
library(readxl)
library(openxlsx)
library(limma)
library(scater)
library(rgl)
library(pca3d)
library(EnhancedVolcano)
library(RColorBrewer)
options(stringsAsFactors = F)
options(dplyr.width = Inf)
getwd()
## not in function
'%nin%' <- Negate('%in%')

# ######## clean memory ######################
# rm(list = ls())
# gc()
# is(dds)
# slotNames(dds)

```

# Normalization
## Background
HIV patients untreated, health control RNASeq data Normalization and QC (quality control). 

Start with the counts table, and compare different normalization methods. DESeq2, TPM, TMM...

DESeq2 is inter-sample comparison normalization method, assuming the majority of the genes are not differentially expressed. 

The top commonly used methods are DESeq (median-of-ratios) and TMM (Trimmed Mean of M values)-edgeR. DESeq and TMM-edgeR were reported to have overall better performance, based on the false positive rate and detection power. 

## Filter Criteria and QC

```{r eda_qc}
# import unnormalized counts table
cnts.raw <- read.delim("All_Sample_geneCounts_raw_counts.txt", header = TRUE, sep = "\t")
cnts.treated.raw <- read.xlsx("HIV-1_infected_HAART_treated_Raw_Counts.xlsx")
head(cnts.raw)
ncol(cnts.raw)
head(cnts.treated.raw)

## explore the gene Symbol "unique"
print("raw data: Symbol")
anyNA(cnts.raw$Symbol)
length(unique(cnts.raw$Symbol)) == nrow(cnts.raw) # there is duplication
gene_sym_sum <- table(cnts.raw$Symbol)
typeof(gene_sym_sum)
sum(gene_sym_sum[gene_sym_sum >= 2]) #  1035 >= 2 symbols
range(gene_sym_sum) # Y_RNA has been used for 490 times

# check unique Gene_ID
print("Gene_ID is unique")
length(unique(cnts.raw$Gene_ID)) == nrow(cnts.raw) # Gene_ID is unique

# generate the common counts table
print("I used Symbol for more information")
cnts <- cnts.raw %>% dplyr::select(-c(Symbol, Length ) ) %>% tibble::column_to_rownames("Gene_ID")
cnts <- as.matrix(cnts)
head(cnts)
dim(cnts)
rna.pid <- colnames(cnts)
# now we have the common counts table 
## pheno
ctrl.id <-  colnames(cnts)[1:13]
ctrl.id
hiv.id <- colnames(cnts)[14:32]
hiv.id
## from dim() we know there are 32 samples
pheno <- data.frame(pid = rna.pid, txt = as.factor(c(rep("Control", 13), 
                                                     rep("HIV", 19)
                                                     )) )
pheno$txt %<>% relevel("Control")
## This is an important step so that DESeq will know to treat the control condition as the reference


# ## without filering
# # using the function from EDASeq
# # using condition here 
# set <- newSeqExpressionSet(as.matrix(round(cnts)),phenoData = data.frame(condition=as.factor(pheno$txt), row.names=colnames(cnts)))
# ##  general QC images  ## 
# 
# ## plotRLE from EDASeq
# plotRLE(set, 
#         outline = FALSE, col=c(rep("Orange", 13), rep("Green", 19)), 
#         main = "Control vs. HIV RLE Plot", 
#         xlab = "Sample", 
#         ylab = "Relative Log Ratio")
# ## PCA plot to show clustering
# ### plotPCA from EDASeq package
# 
# plotPCA(set, col= c(rep("Orange", 13), rep("Green", 19)) )

## add Symbol to the data  
cnts <- as.data.frame(cnts)
cnts$Symbol <- cnts.raw$Symbol
cnts$Length <- cnts.raw$Length
head(cnts)
##filter the raw data and check dim
cnts_fsym <- cnts[rowSums(cnts[, 1:32])>=(5*ncol(cnts[, 1:32])), ]
## should end up around 15 - 20K genes 
ngenes <- nrow(cnts_fsym)
paste("The number of remaining genes: ", ngenes, sep = '')
# all integer in cnts_f
cnts_f <- as.matrix(cnts_fsym[, 1:32])
typeof( as.matrix(cnts_f) )

# using the function from EDASeq
set <- newSeqExpressionSet(as.matrix(cnts_f),phenoData = data.frame(condition=as.factor(pheno$txt), row.names=colnames(cnts_f)))
##  general QC images  ## 
## plotRLE from EDASeq
## boxplots of the log-ratios of the gene-level read counts of each sample to those of a reference sample (defined as the median across the samples). 
EDASeq::plotRLE(set, 
        outline = FALSE, col=c(rep("steelblue", 13), rep("tomato", 19)), 
        main = "Control vs. HIV RLE Plot (With Filtering Before Normalization)", 
        xlab = "Sample", 
        ylab = "Relative Log Ratio")
## PCA plot to show clustering
### plotPCA from EDASeq package

EDASeq::plotPCA(set, col= c(rep("steelblue", 13), rep("tomato", 19)), 
        main = "Control vs. HIV PCA Plot (With Filtering Before Normalization)")
EDASeq::plotPCA(set, col= c(rep("steelblue", 13), rep("tomato", 19)), 
        main = "Control vs. HIV PCA Plot (With Filtering Before Normalization)", k = 3)

## PCA of prcomp
# pc <- princomp(cnts_f, cor=TRUE, scores=TRUE)
# pc$scores
cnts.f.pca <- prcomp(t(cnts_f), center = TRUE, scale. = TRUE, retx = TRUE)
plot(cnts.f.pca)
pca3d(cnts.f.pca, components = 1:3, col= c(rep("steelblue", 13), rep("tomato", 19)),
      title = "Control vs. HIV PCA Plot (With Filtering Before Normalization)",
      radius = 1.5, show.labels = TRUE)
## boxplot 

boxplot(log2(cnts_f), main = "Control vs. HIV Boxplot (With Filtering Before Normalization)",
         xlab = "Sample", ylab = "log2 Counts")




```

## Comparison of DESeq2, TMM and TPM Normalization Methods

```{r normmethods}
################# deseq2 ###########################
# counts from EDASeq (DESeq2)
# pData is phenoData from Biobase
countData <- counts(set) #Matrix with transcripts IDs as rows and sample IDs as columns
colData <- pData(set) #Vector of type list in which the condition column is the treat/control identfier, and the rownames are sample IDs

#Run DESeq function using above objects
print("this is a single factor: condition, and 2 conditions design (2 levels)")

## now using deseq2
dds <- DESeqDataSetFromMatrix(countData = counts(set), colData = pData(set),design = ~ condition)
is(dds)
slotNames(dds)
dds <- estimateSizeFactors(dds)
## normalization factors
sizeFactors(dds)
cnts.deseq2 <- counts(dds, normalized=TRUE)
head(cnts.deseq2)

######## edgeR TMM ###############
## edgeR object
group <- c(rep(1, 13), rep(2, 19))
y <- DGEList(counts= as.matrix(cnts_f), group=group)
## normalization 
y <- calcNormFactors(y, method = "TMM")
y$samples
cnts.edger <- edgeR::cpm(y)
head(cnts.edger)

########## TPM #########
cnts.tpm <- calculateTPM(cnts_f, effective_length = cnts_fsym$Length)
head(cnts.tpm)

######### plots ##################
########### cnts.deseq2
# using the function from EDASeq
# set <- newSeqExpressionSet(cnts.deseq2,phenoData = data.frame(condition=as.factor(pheno$txt), row.names=colnames(cnts_f)))
##  general QC images  ## 
## plotRLE from EDASeq
EDASeq::plotRLE(cnts.deseq2, 
        outline = FALSE, col=c(rep("steelblue", 13), rep("tomato", 19)), 
        main = "Control vs. HIV RLE Plot (With Filtering DESeq2 Normalization)", 
        xlab = "Sample", 
        ylab = "Relative Log Ratio")
## PCA plot to show clustering
### plotPCA from EDASeq package

EDASeq::plotPCA(cnts.deseq2, col= c(rep("steelblue", 13), rep("tomato", 19)), 
        main = "Control vs. HIV PCA Plot (With Filtering DESeq2 Normalization)")
EDASeq::plotPCA(cnts.deseq2, col= c(rep("steelblue", 13), rep("tomato", 19)), 
        main = "Control vs. HIV PCA Plot (With Filtering DESeq2 Normalization)", k = 3)

cnts.deseq2.pca <- prcomp(t(cnts.deseq2), center = TRUE, scale. = TRUE, retx = TRUE)
plot(cnts.deseq2.pca)
pca3d(cnts.deseq2.pca, components = 1:3, col= c(rep("steelblue", 13), rep("tomato", 19)),
      title = "Control vs. HIV PCA Plot (With Filtering DESeq2 Normalization)",
      radius = 1.5, show.labels = TRUE)
pca2d(cnts.deseq2.pca, components = 1:2, col= c(rep("steelblue", 13), rep("tomato", 19)),
      title = "Control vs. HIV PCA Plot (With Filtering DESeq2 Normalization)",
      radius = 1.5)

## boxplot
boxplot(log2(cnts.deseq2), main = "Control vs. HIV Boxplot (With Filtering DESeq2 Normalization)",
         xlab = "Sample", ylab = "log2 Counts")

########### TMM and edgeR
## plotRLE from EDASeq
EDASeq::plotRLE(cnts.edger, 
        outline = FALSE, col=c(rep("steelblue", 13), rep("tomato", 19)), 
        main = "Control vs. HIV RLE Plot (With Filtering TMM (edgeR) Normalization)", 
        xlab = "Sample", 
        ylab = "Relative Log Ratio", ylim = c(-2.1, 2.1))
EDASeq::plotPCA(cnts.edger, col= c(rep("steelblue", 13), rep("tomato", 19)), 
        main = "Control vs. HIV PCA Plot (With Filtering TMM (edgeR) Normalization)")
EDASeq::plotPCA(cnts.edger, col= c(rep("steelblue", 13), rep("tomato", 19)), 
        main = "Control vs. HIV PCA Plot (With Filtering TMM (edgeR) Normalization)", k = 3)

cnts.edger.pca <- prcomp(t(cnts.edger), center = TRUE, scale. = TRUE, retx = TRUE)
plot(cnts.edger.pca)
pca3d(cnts.edger.pca, components = 1:3, col= c(rep("steelblue", 13), rep("tomato", 19)),
      title = "Control vs. HIV PCA Plot (With Filtering TMM Normalization)",
      radius = 1.5, show.labels = TRUE)
pca2d(cnts.edger.pca, components = 1:2, col= c(rep("steelblue", 13), rep("tomato", 19)),
      title = "Control vs. HIV PCA Plot (With Filtering TNN Normalization)",
      radius = 1.5)

boxplot(log2(cnts.edger), main = "Control vs. HIV Boxplot (With Filtering TMM (edgeR) Normalization)",
         xlab = "Sample", ylab = "log2 Counts")

########## TPM
EDASeq::plotRLE(cnts.tpm, 
        outline = FALSE, col=c(rep("steelblue", 13), rep("tomato", 19)), 
        main = "Control vs. HIV RLE Plot (With Filtering TPM Normalization)", 
        xlab = "Sample", 
        ylab = "Relative Log Ratio")
EDASeq::plotPCA(cnts.tpm, col= c(rep("steelblue", 13), rep("tomato", 19)), 
        main = "Control vs. HIV PCA Plot (With Filtering TPM Normalization)")
EDASeq::plotPCA(cnts.tpm, col= c(rep("steelblue", 13), rep("tomato", 19)), 
        main = "Control vs. HIV PCA Plot (With Filtering TPM Normalization)", k = 3 )
boxplot(log2(cnts.tpm), main = "Control vs. HIV Boxplot (With Filtering TPM Normalization)",
         xlab = "Sample", ylab = "log2 Counts")


```

Based on the RLE plots, Boxplots and PCA plots, DESeq2 and TMM of edgeR normalization methods are better. Based on the medians in the RLE plots, the DESeq normalization method was chosen. 
# Differential Expression Analysis 

## DESeq2 Method

The standard differential expression analysis steps in DESeq2 were chosen. The adjusted p value cutoff was set as 0.1 or 0.05. The Benjamini-Hochberg procedure controling false discovery rate (FDR) was used to conduct the p-values adjustment. 

The TMM in edgeR was chosen based on the RLE (relative log expression) plot. We are in favor of more equal spreads across samples. 

```{r deseqDE_edger}
## deseq normalization and DE analysis
register(MulticoreParam(6))
dds <- DESeq(dds)

## the results
res <- results(dds)
summary(res)

## add gene symbol information to the results
res.sym <- res
rownames(res.sym) <- cnts_fsym$Symbol
## res is the result 
resOrdered <- res.sym[order(res.sym$pvalue),]
head(resOrdered[,c(2,6)], 10)

## cutoff 0.1 or 0.05
res.deseq.1 <- resOrdered[resOrdered$padj <= 0.1,]
dim(res.deseq.1)
res.deseq.05 <- resOrdered[resOrdered$padj <= 0.05,]
dim(res.deseq.05)
# minimum padj
res[which.min(res$padj),]
res.sym[which.min(res.sym$padj),]
## top
head(res.deseq.05)

# plots 
## shrinkage of log2 fc to visualize
resultsNames(dds)
resLFC <- lfcShrink(dds, coef="condition_HIV_vs_Control", type="apeglm")
## MA plot
plotMA(resLFC, ylim = c(-5,5))
## plot counts of minimum padj
plotCounts(dds, gene=which.min(res$padj), intgroup="condition",
           main = "EOMES")
plotCounts(dds, gene=which.min(res$padj), intgroup="condition")

## DE analysis of edger
y <- estimateDisp(y)
et <- exactTest(y, pair = 1:2)
dim(et)

### results
topTags(et, n = 10, adjust.method = "BH")
head(et$table)
et$comparison
## summary results
summary(decideTests(et, p.value = 0.05,
            lfc = 0))
paste(1706+ 2540, "genes with FDR 0.05")
summary(decideTests(et, p.value = 0.05,
            lfc = 1))
paste(157 + 680, "genes with FDR 0.05")
summary(decideTests(et, p.value = 0.05,
            lfc = 2))
## 

## counts and results from edgeR
## the counts table from edgeR is cnts.edger
colnames(et$table)
res.edger <- et$table %>% dplyr::mutate(Symbol = cnts_fsym$Symbol,
                                 FDR = p.adjust(et$table$PValue, method = "BH") )
rownames(res.edger) <- rownames(et$table)
head(res.edger)
write.csv(res.edger,
          "~/Documents/gitlab/Cario_RNASeq_Microbiom_Inte/DataProcessed/res.edger.csv",
          row.names = F)
## plot counts, with source code from plotCounts()
group.plot <- factor(c(rep(1,13), rep(2,19)), labels = c("Health Control", "HIV Infected"))
## plot counts
plot(as.integer(group.plot)+ runif(ncol(cnts.edger), -0.1, 0.1), cnts.edger[which.min(res.edger$PValue), ], 
    xlim = c(0.5, length(levels(group.plot)) + 0.8), xaxt="n" ,
    xlab = "Groups", ylab = "Normalized Counts", main = res.edger[which.min(res.edger$PValue), ]$Symbol)
axis(1, at = seq_along(levels(group.plot)), levels(group.plot))
## pos below
text(length(levels(group.plot)) + 0.5, stats::quantile( cnts.edger[which.min(res.edger$PValue), ], 0.95), 
     paste("log2 Fold Change:", round(res.edger[which.min(res.edger$PValue), ]$logFC,2),
             "","Adjusted p-value:",  
           format(res.edger[which.min(res.edger$PValue), ]$FDR,digits=2,scientific=TRUE ),"",
           "Gene_ID: ", rownames(res.edger)[which.min(res.edger$PValue)],
                                           sep = "\n"), pos = 1)
## simple plot counts function
sim_plotcounts <- function(genenumber){
  plot(as.integer(group.plot)+ runif(ncol(cnts.edger), -0.1, 0.1), cnts.edger[genenumber, ], 
    xlim = c(0.5, length(levels(group.plot)) + 0.8), xaxt="n" ,
    xlab = "Groups", ylab = "Normalized Counts", main = res.edger[genenumber, ]$Symbol)
axis(1, at = seq_along(levels(group.plot)), levels(group.plot))
## pos below
text(length(levels(group.plot)) + 0.5, stats::quantile( cnts.edger[genenumber, ], 0.95), 
     paste("log2 Fold Change:", round(res.edger[genenumber, ]$logFC,2),
             "","Adjusted p-value:",  
           format(res.edger[genenumber, ]$FDR,digits=2,scientific=TRUE ),"",
           "Gene_ID: ", rownames(res.edger)[genenumber],
                                           sep = "\n"), pos = 1)
}
## plot top 10
for(i in order(res.edger$PValue)[1:10]){
  sim_plotcounts(i)
}
## check 

```

# Core gene lists

## Core ISGs
Genes that were induced by all the type I interferons tested in vitro.

## IFN-beta specific genes

Genes that were induced specifically by IFN-beta, but not the IFN-alpha subtypes tested

```{r genelists}
# the list of DE genes 
########## by symbol #####
## 4730 out of 19890
sym.deseq.1 <- rownames(res.deseq.1)
## 6585 out of 19890
sym.deseq.05 <- rownames(res.deseq.05)
###### by gene_id
res.order.geneid <- res[order(res$pvalue),]
gid.deseq.1 <- rownames(res.order.geneid[res.order.geneid$padj <= 0.1,])
gid.deseq.05 <- rownames(res.order.geneid[res.order.geneid$padj <= 0.05,])
### summary of results
res.deseq.sum1 <- data.frame(Gene_ID = rownames(res.order.geneid), 
                             Symbol = rownames(resOrdered),
                             Log2FoldChange = resOrdered[,2],
                             pvalue = resOrdered[,5],
                             padj = resOrdered[,6])
kable(head(res.deseq.sum1,10), main = "Top10 Genes Control Vs HIV Untreated",
      digits = c(2,2,2,50,50))


## import pre-defined gene lists 
############## core ISGs ################
isgs <- as.data.frame(read.delim("coreISG"))
dim(isgs)
## DE genes in the core ISGs list 
## directions of regulation 
isgs.de <- base::merge(res.deseq.sum1, isgs, by = "Gene_ID") %>% .[order(.$pvalue), ] %>% 
     dplyr::mutate(Direction = ifelse(.$Log2FoldChange > 0, "Up-regulated", "Down-regulated"))
dim(isgs.de)
##  all symbols are equal, no 0 fold change
sum(isgs.de$Symbol.x != isgs.de$Symbol.y)
sum(isgs.de$Log2FoldChange == 0)
## 230 genes table 
kable(isgs.de[, c(1:5,7)], digits = c(2,2,2,30,30), col.names = 
        c("Gene_ID", "Symbol", "Log2FoldChange", "pvalue", "padj","Direction"))

## FDR 0.05 list 
n.isgs.05 <- nrow(isgs.de[isgs.de$padj <= 0.05, ])
paste("Number of Core ISGs with FDR less than or equal to 0.05", n.isgs.05)


## volcano plot
rownames(isgs.de) <- isgs.de$Symbol.x

EnhancedVolcano(isgs.de,
        lab = rownames(isgs.de),
        x = "Log2FoldChange",
        y = "padj",
        pCutoff = 5e-2,
        FCcutoff = 1,
        pLabellingCutoff = 5e-2,
        ## select labels to show
        # selectLab = c("cg18587484","cg00803922", " cg19425295"),
        ## point and label size 
        transcriptPointSize = 2.0,
        transcriptLabSize = 3.5,
        xlab = bquote(~Log[2]~ "Fold Change"),
        ylab = bquote(~-Log[10]~adjusted~italic(P)),
        title = "Core ISGs: HIV Infected vs Health Control",
        #Modify border and remove gridlines
        gridlines.major = FALSE,
        gridlines.minor = FALSE,
        border = "full",
        borderWidth = 1.0,
        borderColour = "black",
        # the transparence of the dots
        colAlpha = 0.8,
        xlim = c(-4, 4),
        ylim = c(0, -log10(10e-25)),
        # adjust the legend
        legend=c("NS","log2 Fold Change >= 1","adjusted p-value <= 0.05",
            "adjusted p-value <= 0.05 & log2 Fold Change >= 1"),
        legendPosition = "bottom",
        legendLabSize = 9,
        legendIconSize = 3,
        # connectors
        DrawConnectors = TRUE,
        # 
        widthConnectors = 0.3,
        # 
        colConnectors = "grey40",
        col = c("grey30", "forestgreen", "royalblue", "tomato")
        )

EnhancedVolcano(isgs.de,
        lab = rownames(isgs.de),
        x = "Log2FoldChange",
        y = "padj",
        pCutoff = 5e-2,
        FCcutoff = 1.5,
        pLabellingCutoff = 5e-2,
        ## select labels to show
        # selectLab = c("cg18587484","cg00803922", " cg19425295"),
        ## point and label size 
        transcriptPointSize = 2.0,
        transcriptLabSize = 3.5,
        xlab = bquote(~Log[2]~ "Fold Change"),
        ylab = bquote(~-Log[10]~adjusted~italic(P)),
        title = "Core ISGs: HIV Infected vs Health Control",
        #Modify border and remove gridlines
        gridlines.major = FALSE,
        gridlines.minor = FALSE,
        border = "full",
        borderWidth = 1.0,
        borderColour = "black",
        # the transparence of the dots
        colAlpha = 0.8,
        xlim = c(-4, 4),
        ylim = c(0, -log10(10e-25)),
        # adjust the legend
        legend=c("NS","log2 Fold Change >= 1.5","adjusted p-value <= 0.05",
            "adjusted p-value <= 0.05 & log2 Fold Change >= 1.5"),
        legendPosition = "bottom",
        legendLabSize = 9,
        legendIconSize = 3,
        # connectors
        DrawConnectors = TRUE,
        # 
        widthConnectors = 0.3,
        # 
        colConnectors = "grey40",
        col = c("grey30", "forestgreen", "royalblue", "tomato")
        )

######### Beta specific ##############
genesbeta <- as.data.frame(read.delim("genesbeta"))
dim(genesbeta)

## DE genes in the core genesbeta list 
## directions of regulation 
genesbeta.de <- base::merge(res.deseq.sum1, genesbeta, by = "Gene_ID") %>% .[order(.$pvalue), ] %>% 
     dplyr::mutate(Direction = ifelse(.$Log2FoldChange > 0, "Up-regulated", "Down-regulated"))
dim(genesbeta.de)
##  all symbols are equal, no 0 fold change
sum(genesbeta.de$Symbol.x != genesbeta.de$Symbol.y)
genesbeta.de[ which(genesbeta.de$Symbol.x != genesbeta.de$Symbol.y),]
sum(genesbeta.de$Log2FoldChange == 0)
## 423 genes table 
kable(genesbeta.de[, c(1:5,7)], digits = c(2,2,2,30,30), col.names = 
        c("Gene_ID", "Symbol", "Log2FoldChange", "pvalue", "padj","Direction"))

## FDR 0.05 list 
n.genesbeta.05 <- nrow(genesbeta.de[genesbeta.de$padj <= 0.05, ])
paste("Number of Core genesbeta with FDR less than or equal to 0.05", n.genesbeta.05)


## volcano plot
rownames(genesbeta.de) <- genesbeta.de$Symbol.x

EnhancedVolcano(genesbeta.de,
        lab = rownames(genesbeta.de),
        x = "Log2FoldChange",
        y = "padj",
        pCutoff = 5e-2,
        FCcutoff = 1,
        pLabellingCutoff = 5e-2,
        ## select labels to show
        # selectLab = c("cg18587484","cg00803922", " cg19425295"),
        ## point and label size 
        transcriptPointSize = 2.0,
        transcriptLabSize = 3.5,
        xlab = bquote(~Log[2]~ "Fold Change"),
        ylab = bquote(~-Log[10]~adjusted~italic(P)),
        title = "IFN-beta Genes: HIV Infected vs Health Control",
        #Modify border and remove gridlines
        gridlines.major = FALSE,
        gridlines.minor = FALSE,
        border = "full",
        borderWidth = 1.0,
        borderColour = "black",
        # the transparence of the dots
        colAlpha = 0.8,
        xlim = c(-3, 3),
        ylim = c(0, -log10(10e-25)),
        # adjust the legend
        legend=c("NS","log2 Fold Change >= 1","adjusted p-value <= 0.05",
            "adjusted p-value <= 0.05 & log2 Fold Change >= 1"),
        legendPosition = "bottom",
        legendLabSize = 9,
        legendIconSize = 3,
        # connectors
        DrawConnectors = TRUE,
        # 
        widthConnectors = 0.3,
        # 
        colConnectors = "grey40",
        col = c("grey30", "forestgreen", "royalblue", "tomato")
        )


```

### How many of them are significantly altered in the clinical gut biopsies?  (Uninfected vs HIV infected).
For the Core ISGs list, there are `r dim(isgs)[1]`genes, where `r nrow(isgs.de)` of them are in the differential expression list found by DESeq2 with FDR <= 0.1 and `r n.isgs.05` of them are FDR <= 0.05.  

For the IFN-Beta specific genes list, there are `r dim(genesbeta)[1]` genes, where `r nrow(genesbeta.de)` of them are FDR <=0.1, and `r n.genesbeta.05` of them are FDR <= 0.05. 

### Association between genes and clinical outcomes

```{r asso}
## check the counts of gene list
cnts.geneid <- rownames_to_column(data.frame(cnts.deseq2), var = "Gene_ID")
dim(cnts.geneid)
cnts.isgs <- base::merge(cnts.geneid, isgs, by = "Gene_ID") %>% select(-c("Gene_ID")) %>% 
  column_to_rownames(var = "Symbol")
dim(cnts.isgs)
cnts.isgs <- as.matrix(cnts.isgs)
## QC plots
EDASeq::plotRLE(cnts.isgs, 
        outline = FALSE, col=c(rep("steelblue", 13), rep("tomato", 19)), 
        main = "Control vs. HIV RLE Plot (Core ISGs DESeq2 Normalization)", 
        xlab = "Sample", 
        ylab = "Relative Log Ratio")
EDASeq::plotPCA(cnts.isgs, col= c(rep("steelblue", 13), rep("tomato", 19)), 
        main = "Control vs. HIV PCA Plot (Core ISGs DESeq2 Normalization)")
EDASeq::plotPCA(cnts.isgs, col= c(rep("steelblue", 13), rep("tomato", 19)), 
        main = "Control vs. HIV PCA Plot (Core ISGs DESeq2 Normalization)", k = 3 )
boxplot(log2(cnts.isgs), main = "Control vs. HIV Boxplot (Core ISGs DESeq2 Normalization)",
         xlab = "Sample", ylab = "log2 Counts")

## beta specific
# genesbeta
cnts.genesbeta <- base::merge(cnts.geneid, genesbeta, by = "Gene_ID") %>% select(-c("Gene_ID")) %>% 
  column_to_rownames(var = "Symbol")
dim(cnts.genesbeta)
cnts.genesbeta <- as.matrix(cnts.genesbeta)
## QC plots
EDASeq::plotRLE(cnts.genesbeta, 
        outline = FALSE, col=c(rep("steelblue", 13), rep("tomato", 19)), 
        main = "Control vs. HIV RLE Plot (IFN-beta Genes DESeq2 Normalization)", 
        xlab = "Sample", 
        ylab = "Relative Log Ratio")
EDASeq::plotPCA(cnts.genesbeta, col= c(rep("steelblue", 13), rep("tomato", 19)), 
        main = "Control vs. HIV PCA Plot (IFN-beta Genes DESeq2 Normalization)")
EDASeq::plotPCA(cnts.genesbeta, col= c(rep("steelblue", 13), rep("tomato", 19)), 
        main = "Control vs. HIV PCA Plot (IFN-beta Genes DESeq2 Normalization)", k = 3 )
boxplot(log2(cnts.genesbeta), main = "Control vs. HIV Boxplot (IFN-beta Genes DESeq2 Normalization)",
         xlab = "Sample", ylab = "log2 Counts")


# ## Voom transformation
# ## The voom method estimates the mean-variance relationship of the log-counts,
# ## generates a precision weight for each observation and enters these into the limma empirical Bayes analysis pipeline. 
# ## two group design
# design <- model.matrix(~0 + pheno$txt)
# v <- voom(cnts.isgs, design, plot=TRUE)
# v
# voom is more suitable for edgeR


########################### VST Transformation ######################
## DESeq2 has VST, 
## yielding a matrix of values which are now approximately homoskedastic (having constant variance along the range of mean values).
## are useful when checking for outliers or as input for machine learning techniques such as clustering or linear discriminant analysis.

## check size factors
dds$sizeFactor
paste("The size factors vary a lot.")
# ## vst transformation 
# dds <- varianceStabilizingTransformation(dds, blind = FALSE, fitType = "parametric")
# getVarianceStabilizedData(object)
##  If many of genes have large differences in counts due to the experimental design, it is important to set blind=FALSE for downstream analysis.
#################################
## The more the size factors differ, the more residual dependence of the variance on the mean will be found in the transformed data. rlog is a transformation which can perform better in these cases.
rld <- rlog(dds, blind = FALSE, fitType = "parametric")
var.cnts.rld <- assay(rld)
## get linear model ready cnts for core isgs and beta specific
var.rld <- rownames_to_column(data.frame(var.cnts.rld))
colnames(var.rld)[1] <- c("Gene_ID")

## gene level dataset
var.isgs <- base::merge(var.rld, isgs.de, by = "Gene_ID") %>% dplyr::mutate(Symbol = Symbol.x) %>% 
                 select(-c(Symbol.x, Symbol.y))
dim(var.isgs)
var.genesbeta <- base::merge(var.rld, genesbeta.de, by = "Gene_ID") %>% dplyr::mutate(Symbol = Symbol.x) %>% 
                 select(-c(Symbol.x, Symbol.y))
dim(var.genesbeta)


## clustering 
## core isgs
dist.isgs <- dist(t(var.isgs[,2:33]))
plot(hclust(dist.isgs, method = "complete"),
     main = "Cluster Dendrogram by Core ISGs")

## beta specific
dist.genesbeta <- dist(t(var.genesbeta[,2:33]))
plot(hclust(dist.genesbeta, method = "complete"),
     main = "Cluster Dendrogram by IFN-beta Genes")
## kmeans cluster

## kmeans and try k
maxk <- 7
k_c <- 1:maxk
k_sws <- NULL

k_wsm <- NULL
for(i in k_c){
km <- kmeans(t(var.isgs[,2:33]), i, iter.max = 10)
k_sws[i] <- sum(km$withinss)
 if(i == 1){
   k_wsm <- c(km$withinss)
 }else{
   k_wsm <- c(k_wsm, km$withinss)
 }
}
plot(k_c, k_sws, type = "o", 
     xlab = "Number of Clusters by Core ISGs", ylab = "Sum of Within cluster sum of squares")
## kmeans for isgs
kisgs6 <- kmeans(t(var.isgs[,2:33]), 6, iter.max = 10)
kisgs6$cluster
## beta 
for(i in k_c){
km <- kmeans(t(var.genesbeta[,2:33]), i, iter.max = 10)
k_sws[i] <- sum(km$withinss)
 if(i == 1){
   k_wsm <- c(km$withinss)
 }else{
   k_wsm <- c(k_wsm, km$withinss)
 }
}
plot(k_c, k_sws, type = "o", 
     xlab = "Number of Clusters by IFN-Beta Genes", ylab = "Sum of Within cluster sum of squares")
## kmeans for beta
kgbeta3 <- kmeans(t(var.genesbeta[,2:33]), 3, iter.max = 10)
kgbeta3$cluster

################## linear regression ################
## import age, gender and clinical outcomes
## by pid
## basic clinical data
basic_cli_CH <- read.delim("Basic_clinical_control_infected_untreated")
colnames(basic_cli_CH) <- c("pid", "sex", "age", "CD4_Counts", "Plasma_Viral_Load")

# check pid
rna.pid
basic_cli_CH$pid [basic_cli_CH$pid %nin% rna.pid]

#### sample level dataset isgs
isgs.rld.raw <- as.matrix(t(var.isgs[,2:33])) 
colnames(isgs.rld.raw) <- var.isgs$Symbol
## row id to column pid
isgs.rld.raw <- data.frame(isgs.rld.raw)
isgs.rld.raw$pid <- rownames(isgs.rld.raw)
rownames(isgs.rld.raw) <- NULL
## combine gene counts and clinical data 
isgs.rld.lin <- base::merge(isgs.rld.raw, basic_cli_CH, by = "pid")
dim(isgs.rld.lin)

##### sample level dataset genesbeta
genesbeta.rld.raw <- as.matrix(t(var.genesbeta[,2:33])) 
colnames(genesbeta.rld.raw) <- var.genesbeta$Symbol
## row id to column pid
genesbeta.rld.raw <- data.frame(genesbeta.rld.raw)
genesbeta.rld.raw$pid <- rownames(genesbeta.rld.raw)
rownames(genesbeta.rld.raw) <- NULL
## combine gene counts and clinical data 
genesbeta.rld.lin <- base::merge(genesbeta.rld.raw, basic_cli_CH, by = "pid")
dim(genesbeta.rld.lin)

## check same pid
sum(isgs.rld.lin$pid == genesbeta.rld.lin$pid)
# the same order of pid 

########################## linear regression ##############################
## equal length of outcomes and covariates
gene_FunReg <- function(gene_matrix, clinical_variable, genelistname,clin_var_name){
  ## number of gene to test, also the number of multiple test
  n_gene = ncol(gene_matrix)
  ## outcome lm
  outcome_lm = lapply(1:n_gene, function(i){
  lm = lm(gene_matrix[,i]~ clinical_variable + isgs.rld.lin$age + isgs.rld.lin$sex )
  coef = summary(lm)$coefficients[2, ]
  return(coef)
})
   outcome_lm = data.frame(matrix(unlist(outcome_lm), ncol = 4, byrow = TRUE,
                          dimnames = list(
                     c(colnames(gene_matrix)),
                        c("Estimate", "Std.Error", "t.statistic", "p.value"))))

   # adjusted p-value
    outcome_lm =  outcome_lm %>% 
                              dplyr::mutate(FDR = p.adjust(p.value, "BH", n_gene ),
                                               names = colnames(gene_matrix)) %>% 
                                               dplyr::mutate(Estimate = round(Estimate, 10),
                                                      Std.Error = round(Std.Error, 10),
                                                      t.statistic = round(t.statistic,4)
                                                      )%>% 
                                               select(names, everything())
   # sort by p.value
    outcome_lm = outcome_lm[order(outcome_lm$p.value), ]
 
   ## sample size 
    size = length(clinical_variable) - sum(clinical_variable)

   ## summary table 
   kable(outcome_lm, 
         caption = paste("Top Genes from ", genelistname, " Associated with Outcome: ", clin_var_name,
                         " by p.value", " (Sample Size = ", size, ") ", sep = "" , collapse = ""),
         digits = c(2,10,10,4,20,20))
}

############# viral load ####################
## after rlog  the count data to the log2 scale 
## whole isgs 
gene_FunReg(as.matrix(isgs.rld.lin[,2:231]), isgs.rld.lin[,235], "Core ISGs", "Plasma Viral Load")

## betas
gene_FunReg(as.matrix(genesbeta.rld.lin[,2:424]), genesbeta.rld.lin[,428], "IFN-beta Genes", "Plasma Viral Load")


isgs.rld.lin[,235] == genesbeta.rld.lin[,428]
################# CD4 Counts #################
## whole isgs 
gene_FunReg(as.matrix(isgs.rld.lin[,2:231]), isgs.rld.lin[,234], "Core ISGs", "CD4 Counts")

## betas
gene_FunReg(as.matrix(genesbeta.rld.lin[,2:424]), genesbeta.rld.lin[,427], "IFN-beta Genes", "CD4 Counts")

```

### Compare the Associations between Genelists

```{r comparelists}
gene_FunRegRaw <- function(gene_matrix, clinical_variable, genelistname,clin_var_name){
  ## number of gene to test, also the number of multiple test
  n_gene = ncol(gene_matrix)
  ## outcome lm
  outcome_lm = lapply(1:n_gene, function(i){
  lm = lm(gene_matrix[,i]~ clinical_variable + isgs.rld.lin$age + isgs.rld.lin$sex )
  coef = summary(lm)$coefficients[2, ]
  return(coef)
})
   outcome_lm = data.frame(matrix(unlist(outcome_lm), ncol = 4, byrow = TRUE,
                          dimnames = list(
                     c(colnames(gene_matrix)),
                        c("Estimate", "Std.Error", "t.statistic", "p.value"))))

   # adjusted p-value
    outcome_lm =  outcome_lm %>% 
                              dplyr::mutate(FDR = p.adjust(p.value, "BH", n_gene ),
                                               names = colnames(gene_matrix)) %>% 
                                               dplyr::mutate(Estimate = round(Estimate, 10),
                                                      Std.Error = round(Std.Error, 10),
                                                      t.statistic = round(t.statistic,4)
                                                      )%>% 
                                               select(names, everything())
   # sort by p.value
    outcome_lm = outcome_lm[order(outcome_lm$p.value), ]
 
   ## sample size 
    size = length(clinical_variable) - sum(clinical_variable)

   # ## summary table 
   # kable(outcome_lm, 
   #       caption = paste("Top Genes from ", genelistname, " Associated with Outcome: ", clin_var_name,
   #                       " by p.value", " (Sample Size = ", size, ") ", sep = "" , collapse = ""),
   #       digits = c(2,10,10,4,20,20))
    return(outcome_lm)
}

### cd4 counts
################# CD4 Counts #################
## whole isgs 
isgs.cd4 <- gene_FunRegRaw(as.matrix(isgs.rld.lin[,2:231]), isgs.rld.lin[,234], "Core ISGs", "CD4 Counts")

## betas
genesbeta.cd4 <- gene_FunRegRaw(as.matrix(genesbeta.rld.lin[,2:424]), genesbeta.rld.lin[,427], "IFN-beta Genes", "CD4 Counts")

## enchanced volcano plots
## volcano plot
rownames(isgs.cd4) <- isgs.cd4$names
EnhancedVolcano(isgs.cd4,
        lab = rownames(isgs.cd4),
        x = "Estimate",
        y = "FDR",
        FCcutoff = 0.002,
        xlab = bquote("Effect Size"),
        ylab = bquote(~-Log[10]~adjusted~italic(P)),
        title = "Core ISGs: Association with CD4 Counts",
        xlim = c(-0.003, 0.003),
        ylim = c(0, -log10(10e-10)),
        # adjust the legend
        legend=c("NS","log2 Fold Change >= 0.002","adjusted p-value <= 0.05",
            "adjusted p-value <= 0.05 & log2 Fold Change >= 0.002"),
        pLabellingCutoff = 5e-2,
        pCutoff = 5e-2,
        ## select labels to show
        # selectLab = c("cg18587484","cg00803922", " cg19425295"),
        ## point and label size 
        transcriptPointSize = 2.0,
        transcriptLabSize = 3.5,
        #Modify border and remove gridlines
        gridlines.major = FALSE,
        gridlines.minor = FALSE,
        border = "full",
        borderWidth = 1.0,
        borderColour = "black",
        # the transparence of the dots
        colAlpha = 0.8,
        legendPosition = "bottom",
        legendLabSize = 9,
        legendIconSize = 3,
        # connectors
        DrawConnectors = TRUE,
        # 
        widthConnectors = 0.3,
        # 
        colConnectors = "grey40",
        col = c("grey30", "forestgreen", "royalblue", "tomato")
        )

## volcano plot
rownames(genesbeta.cd4) <- genesbeta.cd4$names
EnhancedVolcano(genesbeta.cd4,
        lab = rownames(genesbeta.cd4),
        x = "Estimate",
        y = "FDR",
        FCcutoff = 0.002,
        xlab = bquote("Effect Size"),
        ylab = bquote(~-Log[10]~adjusted~italic(P)),
        title = "IFN-beta Genes: Association with CD4 Counts",
        xlim = c(-0.003, 0.003),
        ylim = c(0, -log10(10e-10)),
        # adjust the legend
        legend=c("NS","log2 Fold Change >= 0.002","adjusted p-value <= 0.05",
            "adjusted p-value <= 0.05 & log2 Fold Change >= 0.002"),
        pLabellingCutoff = 5e-2,
        pCutoff = 5e-2,
        ## select labels to show
        # selectLab = c("cg18587484","cg00803922", " cg19425295"),
        ## point and label size 
        transcriptPointSize = 2.0,
        transcriptLabSize = 3.5,
        #Modify border and remove gridlines
        gridlines.major = FALSE,
        gridlines.minor = FALSE,
        border = "full",
        borderWidth = 1.0,
        borderColour = "black",
        # the transparence of the dots
        colAlpha = 0.8,
        legendPosition = "bottom",
        legendLabSize = 9,
        legendIconSize = 3,
        # connectors
        DrawConnectors = TRUE,
        # 
        widthConnectors = 0.3,
        # 
        colConnectors = "grey40",
        col = c("grey30", "forestgreen", "royalblue", "tomato")
        )

### plot some genes
## plot counts of minimum padj
plotCounts(dds, gene=which.min(res$padj), intgroup="condition",
           main = "EOMES")

```



```{r edgerredo}
## the list of edgeR DE genes 
res.edger.order <- res.edger[order(res.edger$PValue), ]
res.edger.order <- res.edger.order %>% dplyr::mutate(Direction = ifelse(logFC > 0, "Up-regulated", "Down-regulated"))
rownames(res.edger.order) <- rownames(res.edger[order(res.edger$PValue), ])
kable(head(res.edger.order[,-2],10),  digits = c(2,50,2,50,2))
## number of genes
nrow(res.edger.order[res.edger.order$FDR <= 0.05, ])
nrow(res.edger.order[(res.edger.order$FDR <= 0.05) & ((res.edger.order$logFC >= 1) | (res.edger.order$logFC <= -1 )), ])
###
## summary results
summary(decideTests(et, p.value = 0.05,
            lfc = 0))
paste(1706+ 2540, "genes with FDR 0.05")
summary(decideTests(et, p.value = 0.05,
            lfc = 1))
paste(157 + 680, "genes with FDR 0.05")
summary(decideTests(et, p.value = 0.05,
            lfc = 2))
## the results matrix with 0, 1, -1 
# decideTests(et, p.value = 0.05, lfc = 1)
res.edger.05 <- res.edger.order[res.edger.order$FDR <= 0.05, ] 
res.edger.05$Gene_ID <- rownames(res.edger.05)
dim(res.edger.05)

## the two cutoffs
res.edger.05.1 <- res.edger.order[(res.edger.order$FDR <= 0.05) & ((res.edger.order$logFC >= 1) | (res.edger.order$logFC <= -1 )), ]
res.edger.05.1$Gene_ID <- rownames(res.edger.05.1)
dim(res.edger.05.1)
## merge the DE list with predefined genelist 
############################ core ISGs#############################
isgs.edger.05 <- base::merge(res.edger.05, isgs, by = "Gene_ID") 
dim(isgs.edger.05)
write.csv(isgs.edger.05,
             "~/Documents/gitlab/Cario_RNASeq_Microbiom_Inte/DataProcessed/corr/isgs.edger.05.csv")

isgs.edger.05.1 <- base::merge(res.edger.05.1, isgs, by = "Gene_ID") 
dim(isgs.edger.05.1)

colnames(isgs.edger.05 )
## volcano plot
rownames(isgs.edger.05 ) <- isgs.edger.05$Symbol.x

EnhancedVolcano(isgs.edger.05 ,
        lab = rownames(isgs.edger.05 ),
        x = "logFC",
        y = "FDR",
        pCutoff = 5e-2,
        FCcutoff = 1,
        pLabellingCutoff = 5e-2,
        ## select labels to show
        # selectLab = c("cg18587484","cg00803922", " cg19425295"),
        ## point and label size 
        transcriptPointSize = 2.0,
        transcriptLabSize = 3.5,
        xlab = bquote(~Log[2]~ "Fold Change"),
        ylab = bquote(~-Log[10]~adjusted~italic(P)),
        title = "Core ISGs: HIV Infected vs Health Control by edgeR",
        #Modify border and remove gridlines
        gridlines.major = FALSE,
        gridlines.minor = FALSE,
        border = "full",
        borderWidth = 1.0,
        borderColour = "black",
        # the transparence of the dots
        colAlpha = 0.8,
        xlim = c(-4, 4),
        ylim = c(0, -log10(10e-25)),
        # adjust the legend
        legend=c("NS","log2 Fold Change >= 1","adjusted p-value <= 0.05",
            "adjusted p-value <= 0.05 & log2 Fold Change >= 1"),
        legendPosition = "bottom",
        legendLabSize = 9,
        legendIconSize = 3,
        # connectors
        DrawConnectors = TRUE,
        # 
        widthConnectors = 0.3,
        # 
        colConnectors = "grey40",
        col = c("grey30", "forestgreen", "royalblue", "tomato")
        )

############################ ifn beta genes ############################# 
## DE genes in the core genesbeta list 
## directions of regulation 
genesbeta.edger.05 <- base::merge(res.edger.05, genesbeta, by = "Gene_ID") 
dim(genesbeta.edger.05)
write.csv(genesbeta.edger.05,
             "~/Documents/gitlab/Cario_RNASeq_Microbiom_Inte/DataProcessed/corr/genesbeta.edger.05.csv")

genesbeta.edger.05.1 <- base::merge(res.edger.05.1, genesbeta, by = "Gene_ID")
dim(genesbeta.edger.05.1)

########## association ###########
###### still use rlog transformation of deseq2 normalized counts data ###############
## gene level dataset
####### 
rlog.isgs.05 <- base::merge(var.rld, isgs.edger.05, by = "Gene_ID") %>% dplyr::mutate(Symbol = Symbol.x) %>% 
                 select(-c(Symbol.x, Symbol.y))
dim(rlog.isgs.05)

rlog.isgs.05.1 <- base::merge(var.rld, isgs.edger.05.1, by = "Gene_ID") %>% 
  dplyr::mutate(Symbol = Symbol.x) %>% 
                 select(-c(Symbol.x, Symbol.y))
dim(rlog.isgs.05.1)


######### 
rlog.genesbeta.05 <- base::merge(var.rld, genesbeta.edger.05, by = "Gene_ID") %>% 
  dplyr::mutate(Symbol = Symbol.x) %>% 
                 select(-c(Symbol.x, Symbol.y))
dim(rlog.genesbeta.05)

rlog.genesbeta.05.1 <- base::merge(var.rld, genesbeta.edger.05.1, by = "Gene_ID") %>% 
  dplyr::mutate(Symbol = Symbol.x) %>% 
                 select(-c(Symbol.x, Symbol.y))
dim(rlog.genesbeta.05.1)


########### check the effect of regularized log ###############
## compare the effect of rlog
n <- which(rownames(cnts.deseq2) %in% rlog.isgs.05[1, 1])
n
plot(density(as.numeric(  cnts.deseq2[n,] )), main="normalized counts (DESeq2)", cex.main=2)
plot(density(as.numeric(rlog.isgs.05[1,2:33])), main="rlog", cex.main=2)
plot(density(rnorm(10000, 0, 1)), main="ideal", cex.main=2)

## mean and sd relationship
library(vsn)
meanSdPlot(as.matrix(cnts.deseq2), main="normalized counts (DESeq2)")
meanSdPlot(as.matrix(var.cnts.rld), main="rlog")
plot(1:2000, rep(5,2000), type = "l", main = "ideal")

## check number of participants in the association data
nrow(isgs.rld.lin[,2:231])

## make data ready for linear regression
rlog.isgs.05.raw <- as.matrix(t(rlog.isgs.05[,2:33])) 
colnames(rlog.isgs.05.raw) <- rlog.isgs.05$Symbol
## row id to column pid
rlog.isgs.05.raw <- data.frame(rlog.isgs.05.raw)

##  
rlog.isgs.05.1.raw <- as.matrix(t(rlog.isgs.05.1[,2:33])) 
colnames(rlog.isgs.05.1.raw) <- rlog.isgs.05.1$Symbol
## row id to column pid
rlog.isgs.05.1.raw <- data.frame(rlog.isgs.05.1.raw)

######### for genes beta
## make data ready for linear regression
rlog.genesbeta.05.raw <- as.matrix(t(rlog.genesbeta.05[,2:33])) 
colnames(rlog.genesbeta.05.raw) <- rlog.genesbeta.05$Symbol
## row id to column pid
rlog.genesbeta.05.raw <- data.frame(rlog.genesbeta.05.raw)

##  
rlog.genesbeta.05.1.raw <- as.matrix(t(rlog.genesbeta.05.1[,2:33])) 
colnames(rlog.genesbeta.05.1.raw) <- rlog.genesbeta.05.1$Symbol
## row id to column pid
rlog.genesbeta.05.1.raw <- data.frame(rlog.genesbeta.05.1.raw)



## run test association 
############ viral load ################
## whole isgs 
gene_FunReg(as.matrix(rlog.isgs.05.raw), isgs.rld.lin[,235], "Core ISGs", "Plasma Viral Load")
gene_FunReg(as.matrix(rlog.isgs.05.1.raw), isgs.rld.lin[,235], "Core ISGs", "Plasma Viral Load")


## betas
gene_FunReg(as.matrix(rlog.genesbeta.05.raw), isgs.rld.lin[,235], "IFN-beta Genes", "Plasma Viral Load")
gene_FunReg(as.matrix(rlog.genesbeta.05.1.raw), isgs.rld.lin[,235], "IFN-beta Genes", "Plasma Viral Load")


isgs.rld.lin[,235] == genesbeta.rld.lin[,428]
################# CD4 Counts #################
## whole isgs 
gene_FunReg(as.matrix(rlog.isgs.05.raw), isgs.rld.lin[,234], "Core ISGs", "CD4 Counts")
gene_FunReg(as.matrix(rlog.isgs.05.1.raw), isgs.rld.lin[,234], "Core ISGs", "CD4 Counts")

## betas
gene_FunReg(as.matrix(rlog.genesbeta.05.raw), isgs.rld.lin[,234], "IFN-beta Genes", "CD4 Counts")
gene_FunReg(as.matrix(rlog.genesbeta.05.1.raw), isgs.rld.lin[,234], "IFN-beta Genes", "CD4 Counts")


######### scatter plot ######################

######### viral load #####
########## ifn beta
scatter_vl_beta <- function(Symbol){
ggplot(mapping = aes(x = isgs.rld.lin[,235], 
                     y = rlog.genesbeta.05.1.raw[,which(colnames(rlog.genesbeta.05.1.raw) %in% Symbol) ])) + 
  ylab("rlog transformed counts") +
  xlab("Blood HIV Viral Load") +
  geom_point(colour = "gray15", size = 2, shape = 20) +
  geom_smooth(method = "loess", se = FALSE, colour = "steelblue") +
  geom_smooth(method = "lm", se = FALSE, colour = "tomato") +
  theme_bw() +
  theme(
    plot.title = element_text(color="black", size=14, face="bold.italic"),
    axis.title.x = element_text(color="black", size=12, face="bold"),
    axis.title.y = element_text(color="black", size=12, face="bold"),
    axis.text.x = element_text(face="bold", color="black", size=10, angle=0),
    axis.text.y = element_text(face="bold", color="black", size=10, angle=0) ) +
    annotate("text", x = mean(isgs.rld.lin[,234]), 
             y = quantile(rlog.genesbeta.05.1.raw[,which(colnames(rlog.genesbeta.05.1.raw) %in% Symbol) ], 0.95), 
             label = Symbol)

}
for( i in c("THOC3", "AQP3")){
  print(scatter_vl_beta(i))
}
################### cd4 counts #####################
############# core ISGs ############# 
p.irs1 <-  ggplot(mapping = aes(x = isgs.rld.lin[,234], y = rlog.isgs.05.raw[,which(colnames(rlog.isgs.05.raw) %in% "IRS1") ])) + 
  ylab("rlog transformed counts") +
  xlab("Blood CD4 counts") +
  geom_point(colour = "gray15", size = 2, shape = 20) +
  geom_smooth(method = "loess", se = FALSE, colour = "steelblue") +
  geom_smooth(method = "lm", se = FALSE, colour = "tomato") +
  theme_bw() +
  theme(
    plot.title = element_text(color="black", size=14, face="bold.italic"),
    axis.title.x = element_text(color="black", size=12, face="bold"),
    axis.title.y = element_text(color="black", size=12, face="bold"),
    axis.text.x = element_text(face="bold", color="black", size=10, angle=0),
    axis.text.y = element_text(face="bold", color="black", size=10, angle=0) ) +
    annotate("text", x = mean(isgs.rld.lin[,234]), 
             y = quantile(rlog.isgs.05.raw[,which(colnames(rlog.isgs.05.raw) %in% "IRS1") ], 0.95), 
             label = "IRS1")
p.irs1 

scatter_cd4counts_isgs <- function(Symbol){
ggplot(mapping = aes(x = isgs.rld.lin[,234], y = rlog.isgs.05.raw[,which(colnames(rlog.isgs.05.raw) %in% Symbol) ])) + 
  ylab("rlog transformed counts") +
  xlab("Blood CD4 counts") +
  geom_point(colour = "gray15", size = 2, shape = 20) +
  geom_smooth(method = "loess", se = FALSE, colour = "steelblue") +
  geom_smooth(method = "lm", se = FALSE, colour = "tomato") +
  theme_bw() +
  theme(
    plot.title = element_text(color="black", size=14, face="bold.italic"),
    axis.title.x = element_text(color="black", size=12, face="bold"),
    axis.title.y = element_text(color="black", size=12, face="bold"),
    axis.text.x = element_text(face="bold", color="black", size=10, angle=0),
    axis.text.y = element_text(face="bold", color="black", size=10, angle=0) ) +
    annotate("text", x = mean(isgs.rld.lin[,234]), 
             y = quantile(rlog.isgs.05.raw[,which(colnames(rlog.isgs.05.raw) %in% Symbol) ], 0.95), 
             label = Symbol)

}

## plot 
for(i in c("IRS1", "MVB12A", "ODF3B","LGALS3BP", "UBA7", "MYD88","LAG3","TRIM69", "PSMB9", "PSME2")){
  print(scatter_cd4counts_isgs(Symbol = i))
}


## PCA with important genes with TMM NORMARLIZED data 
########## pcd by filtered core ISGs 0.05 log2 FC 1
## PCA of prcomp
# pc <- princomp(cnts_f, cor=TRUE, scores=TRUE)
# pc$scores
cnts.edger.isg.05.1 <- t(cnts.edger[rownames(cnts.edger)  %in% isgs.edger.05.1$Gene_ID, ])
sum(colnames(cnts.edger.isg.05.1 ) != isgs.edger.05.1$Gene_ID)
colnames(cnts.edger.isg.05.1 )  <- isgs.edger.05.1$Symbol.x
edger.pca.isgs.05.1 <- prcomp(cnts.edger.isg.05.1 , 
                     center = TRUE, scale. = TRUE, retx = TRUE)
plot(edger.pca.isgs.05.1)
pca3d(edger.pca.isgs.05.1, components = 1:3, col= c(rep("steelblue", 13), rep("tomato", 19)),
      title = "Core ISGs with FDR 0.05 and log2 FC 1",
      radius = 1.5, show.labels = TRUE)
## 'princomp' can only be used with more units than variables
## pc <- princomp(t(cnts.edger[rownames(cnts.edger)  %in% isgs.edger.05.1$Gene_ID, ]), cor=TRUE, scores=TRUE)
getwd()
######## key genes
###### LAG3 ODF3B
num.keygenes <- which( rownames(edger.pca.isgs.05.1$rotation) %in% c("LAG3", "ODF3B","NLRC5") )
plot3d(edger.pca.isgs.05.1$x[,1:3], col= c(rep("steelblue", 13), rep("tomato", 19)) , 
       size = 1, type = 's')
text3d(edger.pca.isgs.05.1$rotation[num.keygenes,1:3]*20, 
       texts=rownames(edger.pca.isgs.05.1$rotation)[num.keygenes], col="black", justify = "left",
       font = 5)
coords <- NULL
for (i in num.keygenes) {
  coords <- rbind(coords, rbind(c(0,0,0),edger.pca.isgs.05.1$rotation[i,1:3]*20))
}
lines3d(coords, col="black", lwd=2)

# # add legend
# legend3d("topright", legend = paste( c('Health Control', 
#                                               'HIV Infected')), 
#          pch = 16, col = c("steelblue", "tomato"), cex=1, inset=c(0.02))
# 
# # capture snapshot
# # snapshot3d(filename = '3dplot.png', fmt = 'png')
# 
# ######### snapshot
# rgl.bringtotop()
# rgl.viewpoint(0, 20)
# rgl.snapshot("test.png")

```

```{r focusTMM}
## # all integer in cnts_f, filtered counts
### cnts  cnts.edger cnts.deseq2 cnts.tpm

## var.cnts.rld rlog transformed counts, based on DESeq2 normalization

## pca

## edger.pca.isgs.05.1$x  the matrix define the scatter of points, 1 to 13 health control

##  explore scater
# the counts data
# data("sc_example_counts")
# 
# ## the dataframe has condition information
# data("sc_example_cell_info")
# 
# example_sce <- SingleCellExperiment(
#                  assays = list(counts = sc_example_counts),
#                  colData = sc_example_cell_info)

rna_info <- data.frame(pid = rna.pid, Group = as.factor(c(rep("Control", 13), 
                                                     rep("HIV Infected", 19)
                                                     )) )
rna_info$Group %<>% relevel("Control")
## cnts. list
cnts.list <- list(cnts_f, cnts.edger, cnts.deseq2, cnts.tpm)
cnts.names <- c("Filtered Counts", "TMM Normalized Counts",
                "DESeq2 Normalized Counts", "TPM Normalized Counts")
for (i in 1:4){
  j  = cnts.list[[i]]
  ## using scater, start with an object
example_sce <- SingleCellExperiment(
                 assays = list(counts = j),
                 colData = rna_info)
## RLE plots
print( scater::plotRLE(example_sce, exprs_values = "counts", exprs_logged = FALSE,
        colour_by = "Group", style = "minimal") + scale_x_discrete("Samples", labels = rna.pid) + labs(caption = paste("RLE Plot:" ,cnts.names[i]) ) )
## pca plots
example_sce <- SingleCellExperiment(
                 assays = list(logcounts = j),
                 colData = rna_info)
print( scater::plotPCA(example_sce, colour_by = "Group", size_by = "Group") + labs(caption = paste(  "PCA Plot:",cnts.names[i]) ))
}
## define the gene list 
invitro_genes <- c("Mx2", "APOBEC3G", "Tetherin", "ISG15", "CD169")
## turn in to upper case
invitro_genes[ sapply(invitro_genes, toupper) %nin% res.edger.05$Symbol ]
invitro_genes[ sapply(invitro_genes, toupper) %nin% res.edger.05.1$Symbol ]

## check
# "CD169" %in% res.edger.05$Symbol
# "TETHERIN" %in% res.edger.05$Symbol
# "CD169" %in% res.edger.order$Symbol
# "TETHERIN" %in% res.edger.order$Symbol
"CD169" %in% cnts.raw$Symbol
"TETHERIN" %in% cnts.raw$Symbol
# "CD169" %in% isgs$Symbol
# "TETHERIN" %in% isgs$Symbol
# "CD169" %in% genesbeta$Symbol
# "TETHERIN" %in% genesbeta$Symbol
## 
cnts.raw[ cnts.raw$Symbol %in% "APOBEC3G", ]
## a fold change is the ratio of the averages
vitro_table_seq <- res.edger.05[res.edger.05$Symbol %in% sapply(invitro_genes, toupper),c(4,1,5)]
vitro_table_seq[4,] <- c("Tetherin", "", "")
vitro_table_seq[5,] <- c("CD169", "", "")
colnames(vitro_table_seq) <- c("Symbol", "RNA-Seq logFC","FDR")
vitro_table_seq
############## RECHECK vitro genes#########################
## read in all clinical data
clinical_ready <- read_excel("~/Documents/gitlab/Cario_RNASeq_Microbiom_Inte/DataProcessed/clinical_ready.xlsx")
## clinical_ready <- unlist(clinical_ready)
clinical_ready <- as.data.frame(clinical_ready)
# factor(clinical_ready$Group)
# clinical_ready$Group %<>% relevel("control")
for(j in 1:5){
i = c(22,20,19,21,23)[j]
t_vitro = t.test(clinical_ready[,i] ~ clinical_ready$Group)
vitro_table_seq[j,4] <-round(log2(t_vitro$estimate[2] - t_vitro$estimate[1]),2)
vitro_table_seq[j,5] <- round(t_vitro$p.value,5)
colnames(vitro_table_seq)[c(4,5)] <- c("qRT-PCR logFC", "t-test p.value")
}
kable(vitro_table_seq, digits = c(2,2,2,2,5))
# formatC(numb, format = "e", digits = 2)
## format(res.edger[which.min(res.edger$PValue), ]$FDR,digits=2,scientific=TRUE )
## formatC
## prettyNum()

############## get dataet ready for association test ###########
dim(clinical_ready)
clinical_ready$pid[1:24] <- gsub("MIHIV", "H", clinical_ready$pid[1:24]) 
clinical_ready$pid[25:38] <- gsub("MIHIV", "C", clinical_ready$pid[25:38]) 
clinical <- clinical_ready[clinical_ready$pid %in% rna.pid, ]
dim(clinical)
## now we have data clinical
## using FDR 0.05 cutoff to get the rlog data ready
## isgs.edger.05 has the gene symbol list 
length(unique(isgs$Symbol)) == nrow(isgs)  ## every symbol in isgs is unique
length(unique(genesbeta$Symbol)) == nrow(genesbeta) 
nrow(isgs); nrow(genesbeta) 
length(unique(res.edger$Symbol)) == nrow(res.edger)
## res.edger
sum(res.edger$Symbol != cnts_fsym$Symbol )
paste("Using Symbol as identifier")
  ######### isgs #############
isgs.edger <- res.edger[ res.edger$Symbol %in% isgs$Symbol, ]
nrow(isgs.edger) == nrow(isgs)
## 117 genes out of 230 FDR 0.05 in ISGs.
nrow(isgs.edger[ isgs.edger$FDR <= 0.05, ]  )
## ranked by logFC
isgs.save <- isgs.edger[base::order(isgs.edger$logFC, decreasing = TRUE), c(4,1,5)]
isgs.save$Gene_ID <- row.names(isgs.save)

write.xlsx(isgs.save,
           "~/Documents/gitlab/Cario_RNASeq_Microbiom_Inte/DataProcessed/ISGs_DEanalysis_results.xlsx",
           sheetName="ISGs Results")

  ########  genes beta ##############
genesbeta.edger <- res.edger[ res.edger$Symbol %in% genesbeta$Symbol, ]
genesbeta$Symbol [genesbeta$Symbol %nin%res.edger$Symbol]
genesbeta.edger[423,] <- res.edger[row.names(res.edger) ==  genesbeta$Gene_ID [genesbeta$Symbol %nin%res.edger$Symbol] , ]
nrow(genesbeta.edger) == nrow(genesbeta)
# 130 genes out of 423 FDR IFN-beta genes.
genesbeta.edger[423,4] <- genesbeta$Symbol [genesbeta$Symbol %nin%res.edger$Symbol]
nrow(isgs.edger[ genesbeta.edger$FDR <= 0.05, ]  )

genesbeta.save <- genesbeta.edger[base::order(genesbeta.edger$logFC, decreasing = TRUE), c(4,1,5)]
genesbeta.save$Gene_ID <- row.names(genesbeta.save)

write.xlsx(genesbeta.save,
           "~/Documents/gitlab/Cario_RNASeq_Microbiom_Inte/DataProcessed/IFNbeta_DEanalysis_results.xlsx",
           sheetName="IFNbeta Results")


### by the way, save data file
## rlog transformed data
sum(var.rld$Gene_ID != row.names(cnts_fsym))
var.rld.save <- var.rld %>% dplyr::mutate(Symbol = cnts_fsym$Symbol) %>% dplyr::select(Gene_ID, Symbol, everything())

write.xlsx(var.rld.save,
           "~/Documents/gitlab/Cario_RNASeq_Microbiom_Inte/DataProcessed/rlog_counts_linear_regression.xlsx",
           sheetName="rlog transformed counts")

## TMM normalized counts
sum(row.names(cnts.edger) != row.names(cnts_fsym))
cnts.edger <- data.frame(cnts.edger) %>% dplyr::mutate(Symbol = cnts_fsym$Symbol,
                                                       Gene_ID = row.names(cnts_fsym))  %>% dplyr::select(Gene_ID, Symbol, everything())
write.xlsx(cnts.edger,
           "~/Documents/gitlab/Cario_RNASeq_Microbiom_Inte/DataProcessed/TMM_normalized_counts.xlsx",
           sheetName="TMM Normalized Counts")

## ## gene level dataset
var.isgs.05 <- base::merge(var.rld, isgs.edger.05, by = "Gene_ID") %>% dplyr::mutate(Symbol = Symbol.x) %>% 
                 select(-c(Symbol.x, Symbol.y))

dim(var.isgs.05)
isgs.05.lin <- t(var.isgs.05[2:33])
base::colnames(isgs.05.lin) <- var.isgs.05$Symbol

head(isgs.05.lin)
write.csv(isgs.05.lin,
             "~/Documents/gitlab/Cario_RNASeq_Microbiom_Inte/DataProcessed/corr/isgs.05.lin.csv")
           
var.genesbeta.05 <- base::merge(var.rld, genesbeta.edger.05, by = "Gene_ID") %>% dplyr::mutate(Symbol = Symbol.x) %>% select(-c(Symbol.x, Symbol.y))
dim(var.genesbeta.05)

genesbeta.05.lin <- t(var.genesbeta.05[2:33])
base::colnames(genesbeta.05.lin) <- var.genesbeta.05$Symbol
head(genesbeta.05.lin)
write.csv(genesbeta.05.lin,
             "~/Documents/gitlab/Cario_RNASeq_Microbiom_Inte/DataProcessed/corr/genesbeta.05.lin.csv")

################### make clinical as the same pid order of gene #################
## not the same order
## check
sum(as.matrix(rlog.isgs.05.raw) != isgs.05.lin)
sum(isgs.rld.lin[,234] != clinical[,6])
sum(isgs.rld.lin$age == clinical$age)
## 
isgs.05.lin.order <- data.frame(isgs.05.lin) %>% dplyr::mutate(pid = row.names(isgs.05.lin))
clinical_order <- base::merge(clinical, isgs.05.lin.order, by = "pid")
clinical_order <- clinical_order[,c(1:18,24)]
write.csv(clinical_order,
             "~/Documents/gitlab/Cario_RNASeq_Microbiom_Inte/DataProcessed/corr/clinical_order.csv")

## clinical_order$pid == row.names(isgs.05.lin)

########### association ################
## data isgs.05.lin genesbeta.05.lin and clinical 

########################## linear regression ##############################
## equal length of outcomes and covariates
gene_IFNReg <- function(gene_matrix, clinical_variable, clin_var_name){
  # get names ready
  genelistname = base::colnames(gene_matrix)
  ## number of gene to test, also the number of multiple test
  n_gene = ncol(gene_matrix)
  ## outcome lm
  outcome_lm = lapply(1:n_gene, function(i){
  lm = lm(gene_matrix[,i]~ clinical_variable + clinical_order$age + clinical_order$sex )
  coef = summary(lm)$coefficients[2, ]
  return(coef)
})
   outcome_lm = data.frame(matrix(unlist(outcome_lm), ncol = 4, byrow = TRUE,
                          dimnames = list(
                     c(colnames(gene_matrix)),
                        c("Estimate", "Std.Error", "t.statistic", "p.value"))))

   # adjusted p-value
    outcome_lm =  outcome_lm %>% 
                              dplyr::mutate(FDR = p.adjust(p.value, "BH", n_gene ),
                                               names = colnames(gene_matrix)) %>% 
                                               dplyr::mutate(Estimate = round(Estimate, 10),
                                                      Std.Error = round(Std.Error, 10),
                                                      t.statistic = round(t.statistic,4)
                                                      )%>% 
                                               select(names, everything())
   # sort by p.value
    outcome_lm = outcome_lm[order(outcome_lm$p.value), ]
 
   ## sample size 
    size = sum(!is.na(clinical_variable))

   ## summary table 
   return(list(results = data.frame(outcome_lm), size = size, clinical = clin_var_name ))
}

clinical_names <- c("Blood CD4 T Cell Counts (cells/ul)", "Plasma Viral Load", "Tissue HIV RNA (per CD4 T cell)",
                    "Tissue CD4 T Cell Counts (number/g)", "IL-6 (pg/ml)", "CRP (ug/ml)", "iFABP (pg/ml)",
                    "sCD27 (U/ml)", "CD14 (ng/ml)", "LPS (pg/ml)", "LTA (OD)", 
                    base::paste("IFN", '\u03b1', sep = "" ),  base::paste("IFN", '\u03b2', sep = "" ))
n_clinical <- length(clinical_names)
n_clinical
clinical_sum <- matrix(NA, 13,8)
for(i in 1:n_clinical) {
  ## number of clinical virable
  j = c(6:18)[i]
  cliname = base::colnames(clinical_order)[j]
  ## linear regression
  lin_res_isgs = gene_IFNReg(isgs.05.lin, clinical_order[,j], clinical_names[i])
  lin_res_genesbeta = gene_IFNReg(genesbeta.05.lin, clinical_order[,j], clinical_names[i])
  ## save data
 
  write.xlsx(lin_res_isgs$results,
            paste("~/Documents/gitlab/Cario_RNASeq_Microbiom_Inte/DataProcessed/isgs_", cliname,".xlsx", sep = ""),
           sheetName= paste("ISGs_", cliname, sep = "") )
  write.xlsx(lin_res_genesbeta$results,
            paste("~/Documents/gitlab/Cario_RNASeq_Microbiom_Inte/DataProcessed/ifnbeta_", cliname,".xlsx", sep = ""),
           sheetName= paste("ifnbeta_", cliname, sep = ""))
  
  ## check basic
  if(lin_res_isgs$size == lin_res_genesbeta$size){
    print("Good")
  }else stop("sample size wrong")
  if(lin_res_isgs$clinical == lin_res_genesbeta$clinical){
    print("Good")
  }else stop("clinical parameter wrong")
  ## number of genes
  nisgs = nrow(lin_res_isgs$results)
  ngenesbeta = nrow(lin_res_genesbeta$results)
  ################## summary table ###################
  ###  sig prop
  isgs.sig.no =  sum(lin_res_isgs$results$FDR <= 0.05)
  genesbeta.sig.no =  sum(lin_res_genesbeta$results$FDR <= 0.05)
  if( (isgs.sig.no >= 5) & (genesbeta.sig.no >= 5) ){
    prop.test.sig = prop.test(x = c(isgs.sig.no, genesbeta.sig.no), n = c(nisgs, ngenesbeta), correct = FALSE)
    sig.p = prop.test.sig$p.value
  }else{
    prop.test.sig = prop.test(x = c(isgs.sig.no, genesbeta.sig.no), n = c(nisgs, ngenesbeta), correct = TRUE)
    sig.p = prop.test.sig$p.value
  }
         ########### volcano plots ################
  if(min(lin_res_isgs$results$FDR) <= 0.05){
    rownames(lin_res_isgs$results) <- lin_res_isgs$results$names
  p1 <- EnhancedVolcano(lin_res_isgs$results ,
        lab = rownames(lin_res_isgs$results),
        x = "Estimate",
        y = "p.value",
        pCutoff = ifelse( min(lin_res_isgs$results$FDR) <= 0.05,
                          lin_res_isgs$results$p.value[max(which(lin_res_isgs$results$FDR <= 0.05))] + 0.005, 
                          0.05),
        FCcutoff = round(max(abs(quantile(lin_res_isgs$results$Estimate, c(0.25, 0.75) ))),1),
        title = paste("Core ISGs: association with ", lin_res_genesbeta$clinical , sep = ""),
        xlim = range(lin_res_isgs$results$Estimate)*1.3,
        ylim = c(0, -log10(min(lin_res_isgs$results$p.value)/5000)),
        legend=c("NS","Slope","FDR Significant",
            "FDR Significant & Slope"),
        ## select labels to show
        # selectLab = c("cg18587484","cg00803922", " cg19425295"),
        ## point and label size 
        transcriptPointSize = 2,
        transcriptLabSize = 3,
        xlab = bquote(~Log[2]~ "Slope"),
        ylab = bquote(~-Log[10]~italic(P)),
        
        #Modify border and remove gridlines
        gridlines.major = FALSE,
        gridlines.minor = FALSE,
        border = "full",
        borderWidth = 1.0,
        borderColour = "black",
        # the transparence of the dots
        colAlpha = 0.8,
   
        # adjust the legend
        legendPosition = "bottom",
        legendLabSize = 9,
        legendIconSize = 3,
        # connectors
        DrawConnectors = TRUE,
        # 
        widthConnectors = 0.2,
        # 
        colConnectors = "grey40",
        col = c("grey30", "forestgreen", "royalblue", "tomato")
        )
  ggsave(paste("Volcanol_Plot_ISGs_", cliname, sep = ""), width = 6, height = 9, device = "tiff", path = "~/Documents/gitlab/Cario_RNASeq_Microbiom_Inte/Reports/plots", dpi = 300)
  } else{print("No Sig.")}
 
  if(min(lin_res_genesbeta$results$FDR) <= 0.05){
  rownames(lin_res_genesbeta$results) <- lin_res_genesbeta$results$names
  p1 <- EnhancedVolcano(lin_res_genesbeta$results ,
        lab = rownames(lin_res_genesbeta$results),
        x = "Estimate",
        y = "p.value",
        pCutoff = ifelse( min(lin_res_genesbeta$results$FDR) <= 0.05,
                          lin_res_genesbeta$results$p.value[max(which(lin_res_genesbeta$results$FDR <= 0.05))] + 0.005, 
                          0.005),
        FCcutoff = max(abs(quantile(lin_res_genesbeta$results$Estimate, c(0.25, 0.75) ))) ,
        title = paste("IFNBeta Genes: association with ", lin_res_genesbeta$clinical , sep = ""),
        xlim = range(lin_res_genesbeta$results$Estimate)*1.3,
        ylim = c(0, -log10(min(lin_res_genesbeta$results$p.value)/5000)),
        legend=c("NS","Slope","FDR Significant",
            "FDR Significant & Slope"),
        ## select labels to show
        # selectLab = c("cg18587484","cg00803922", " cg19425295"),
        ## point and label size 
        transcriptPointSize = 2,
        transcriptLabSize = 3,
        xlab = bquote("Slope"),
        ylab = bquote(~-Log[10]~italic(P)),
        
        #Modify border and remove gridlines
        gridlines.major = FALSE,
        gridlines.minor = FALSE,
        border = "full",
        borderWidth = 1.0,
        borderColour = "black",
        # the transparence of the dots
        colAlpha = 0.8,
   
        # adjust the legend
        legendPosition = "bottom",
        legendLabSize = 9,
        legendIconSize = 3,
        # connectors
        DrawConnectors = TRUE,
        # 
        widthConnectors = 0.2,
        # 
        colConnectors = "grey40",
        col = c("grey30", "forestgreen", "royalblue", "tomato")
        )
  ggsave(paste("Volcanol_Plot_IFNBeta_", cliname, sep = ""), width = 6, height = 9, device = "tiff", path = "~/Documents/gitlab/Cario_RNASeq_Microbiom_Inte/Reports/plots", dpi = 300)
  } else{print("No Sig.")}
 
 
  
  ## positive prop
  isgs.pos.no = sum(lin_res_isgs$results$Estimate > 0)
  genesbeta.pos.no = sum(lin_res_genesbeta$results$Estimate > 0)
   if( (isgs.pos.no >= 5) & (genesbeta.pos.no >= 5) ){
    prop.test.pos = prop.test(x = c(isgs.pos.no, genesbeta.pos.no), n = c(nisgs, ngenesbeta), correct = FALSE)
    pos.p = prop.test.pos$p.value
  }else{
    prop.test.pos = prop.test(x = c(isgs.pos.no, genesbeta.pos.no), n = c(nisgs, ngenesbeta), correct = TRUE)
    pos.p = prop.test.pos$p.value
  }
  ## If you check prop.test, chisq.test and z-test on your data then they all give you the same p-value
  ## continuity correction if anyone less than 5
  ## the clinical summary table 
  clinical_sum[i, ] <- c(lin_res_genesbeta$clinical, lin_res_genesbeta$size,
                         isgs.sig.no/nisgs, genesbeta.sig.no/ngenesbeta , sig.p, 
                         isgs.pos.no/nisgs,
                         genesbeta.pos.no/ngenesbeta , pos.p)
  
  colnames(clinical_sum) <- c("Clinical Parameter", "Sample Size", "ISGs: Sig. Prop.",
                              "IFNbeta Genes: Sig.", "p value",
                              "ISGs: Positive Corr. Prop.",
                              "IFNbeta Genes: Pos.", "p value (Pos)")
}


## clinical_sum
write.xlsx(clinical_sum, "~/Documents/gitlab/Cario_RNASeq_Microbiom_Inte/DataProcessed/clinical_summary.xlsx")
clinical_sum[,c(3,4,6,7)] <- round(as.numeric( as.character( clinical_sum[,c(3,4,6,7) ]) ), 4)
clinical_sum[,c(5,8)] <- format(as.numeric( as.character( clinical_sum[,c(5,8) ]) ), digits = 3, scientific = TRUE)
clinical_sum[12:13,1] <- c("IFNalpha", "IFNbeta")
kable(data.frame(clinical_sum[,c(1:5)]), digits = c(2,2,4,4,40))
kable(data.frame(clinical_sum[,c(1,2,6:8)]), digits = c(2,2,4,4,40))

```


# Reference
1. Biomed Res Int.2015 Jun 15. doi: 10.1155/2015/621690, The Impact of Normalization Methods on RNA-Seq Data Analysis, J. Zyprych-Walczak, A. Szabelska, L. Handschuh, K. Grczak, K. Klamecka, M. Figlerowicz, and I. Siatkowski

2. A comparison of per sample global scaling and per gene normalization methods for differential expression analysis of RNA-seq data, PLos One. 2017. 

3. Voom: precision weights unlock linear model analysis tools for RNA-seq read counts, Charity W Law, Yunshun Chen, Wei Shi and Gordon K Smyth, Genome Biology, 2014 15:R29. 

4. Michael I Love, Wolfgang Huber, Simon Anders: Moderated estimation of fold change and dispersion for RNA-seq data with DESeq2. Genome Biology 2014, 15:550. http://dx.doi.org/10.1186/s13059-014-0550-8.




