---
title: "HIV Treated vs Control"
author: "Guannan Shen"
date: "April 28, 2019"
output: 
  html_document:
    number_sections: yes
    theme: united
    toc: yes
    toc_depth: 5
    toc_float: yes
  word_document:
    toc: yes
    toc_depth: '5'
  pdf_document:
    latex_engine: lualatex
    number_sections: yes
    toc: yes
    toc_depth: 5
---

```{r setup, include=FALSE, cache = FALSE}
require("knitr")
opts_chunk$set(tidy.opts=list(width.cutoff=60),tidy=TRUE)
opts_chunk$set(engine = "R")
knitr::opts_chunk$set(echo = T)
knitr::opts_chunk$set(message = F)
knitr::opts_chunk$set(warning = F)
## setting wd in DELL
## opts_knit$set(root.dir = "~/Stats/CIDA_OMICs/CIDA_OMICS/7659Stats_Genetics/HW5/")
## setting working directory in asus 
## opts_knit$set(root.dir = "C:/Users/hithr/Documents/Stats/gitlab/Cario_RNASeq_Microbiom_Inte/DataRaw/") 
## setting working directory in ubuntu
opts_knit$set(root.dir = "~/Documents/gitlab/Cario_RNASeq_Microbiom_Inte/DataRaw/interferome/")
                                                 
## cache = F, if cache = T, will not revaluate code chunk everytime
## double or more space to insert a line break

```



```{r libs, warning=FALSE, echo=F}
######################
## Set up workspace
######################
rm(list = ls())
library(edgeR)
library(EDASeq)
library(DESeq2)
library(knitr)
library(tidyverse)
library(magrittr)
library(stats)
library(BiocParallel)
library(readxl)
library(openxlsx)
library(limma)
library(scater)
library(rgl)
library(pca3d)
library(EnhancedVolcano)
library(RColorBrewer)
options(stringsAsFactors = F)
options(dplyr.width = Inf)
getwd()
## not in function
'%nin%' <- Negate('%in%')

# ######## clean memory ######################
# rm(list = ls())
# gc()
# is(dds)
# slotNames(dds)

```

# Data Preprocessing
15 HIV-infected-treated donors vs. 11 healthy controls.  

Filtering was applied. The inclusion criteria is the gene must have a total counts (sum up all samples) more than or equal to the number of samples times 5.  


```{r eda_qc, echo=F}
# import data
cnts.raw <- read.xlsx("HIV-1_infected_HAART_treated_Raw_Counts.xlsx")

########### need to realign the samples ###################

# filtering, the same 
## filter the raw data and check dim
cnts_fsym <- cnts.raw[rowSums(cnts.raw[, 4:29])>=(5*ncol(cnts.raw[, 4:29])), ]
dim(cnts_fsym)
row.names(cnts_fsym) <- NULL

### get matrix
cnts_f <- cnts_fsym %>% dplyr::select(-c(Symbol, Length ) ) %>% 
            tibble::column_to_rownames("Gene_ID") %>% as.matrix()


########### pheno grouping ####################
dim(cnts_f)
ctrl.id <-  colnames(cnts_f)[1:11]
hiv.id <- colnames(cnts_f)[12:26]

rna.pid <- colnames(cnts_f)
pheno <- data.frame(pid = rna.pid, txt = as.factor(c(rep("Control", 11), 
                                                     rep("HIV-Treated", 15)
                                                     )) )
pheno$txt %<>% relevel("Control")

############ EDA ############
# using the function from EDASeq
set <- newSeqExpressionSet(as.matrix(cnts_f),phenoData = data.frame(condition=as.factor(pheno$txt), row.names=colnames(cnts_f)))

################# deseq2 ###########################
## now using deseq2
dds <- DESeqDataSetFromMatrix(countData = counts(set), colData = pData(set),design = ~ condition)
dds <- estimateSizeFactors(dds)
## normalization factors
paste("size Factors of DESeq2")
sizeFactors(dds)
cnts.deseq2 <- counts(dds, normalized=TRUE)

######## edgeR TMM ###############
## edgeR object
group <- c(rep(1, 11), rep(2, 15))
y <- DGEList(counts= as.matrix(cnts_f), group=group)
## normalization 
y <- calcNormFactors(y, method = "TMM")
paste("Scaling Factors of edgeR")
y$samples
cnts.edger <- edgeR::cpm(y)

########## TPM #########
cnts.tpm <- calculateTPM(cnts_f, effective_length = cnts_fsym$Length)

###########3 using scater for QC EDA ##################
rna_info <- data.frame(pid = rna.pid, Group = as.factor(c(rep("Control", 11), 
                                                     rep("HIV-Treated", 15)
                                                     )) )
rna_info$Group %<>% relevel("Control")

## cnts. list
cnts.list <- list(cnts_f, cnts.edger, cnts.deseq2, cnts.tpm)
cnts.names <- c("Filtered Counts", "TMM Normalized Counts",
                "DESeq2 Normalized Counts", "TPM Normalized Counts")
par(mfrow = c(1,2))
for (i in 1:4){
  j  = cnts.list[[i]]
  ## using scater, start with an object
example_sce <- SingleCellExperiment(
                 assays = list(counts = j),
                 colData = rna_info)
## RLE plots
print( scater::plotRLE(example_sce, exprs_values = "counts", exprs_logged = FALSE,
        colour_by = "Group", style = "minimal") + scale_x_discrete("Samples", labels = rna.pid) + labs(caption = paste("RLE Plot:" ,cnts.names[i]) ) )
## pca plots
example_sce <- SingleCellExperiment(
                 assays = list(logcounts = j),
                 colData = rna_info)
print( scater::plotPCA(example_sce, colour_by = "Group", size_by = "Group") + labs(caption = paste(  "PCA Plot:",cnts.names[i]) ))
}

######### the gene list is a fair list###########
## define the gene list 
invitro_genes <- c("Mx2", "APOBEC3G", "Siglec1", "ISG15", "Bst2")
## turn in to upper case
invitro_genes[ sapply(invitro_genes, toupper) %nin% cnts_fsym$Symbol ]

```

From the scaling factor, we can tell the between-samples variance is samller compared with the HIV-untreated dataset. For the previous dataset, the more the size factors differ, the more residual dependence of the variance on the mean will be found in the transformed data. **rlog** is a transformation which can perform better in these cases.  

For this dataset, a variance stabilizing transformation, **vst** should be more proper.  

varianceStabilizingTransformation(object, blind = FALSE, fitType = "parametric")  

getVarianceStabilizedData(object)

```{r de_edger}
# make sure y is the object from edger
## DE analysis of edger
y <- estimateDisp(y)
et <- exactTest(y, pair = 1:2)
dim(et)

### results
topTags(et, n = 10, adjust.method = "BH")
head(et$table)
et$comparison
## summary results
summary(decideTests(et, p.value = 0.05,
            lfc = 0))
# paste(1706+ 2540, "genes with FDR 0.05")

summary(decideTests(et, p.value = 0.05,
            lfc = 1))
# paste(157 + 680, "genes with FDR 0.05")

summary(decideTests(et, p.value = 0.05,
            lfc = 2))
## 

```


