---
title: "PCA Gene Lists Untreated"
author: "Guannan Shen"
date: "April 29, 2019"
output: 
  html_document:
    number_sections: yes
    theme: united
    toc: yes
    toc_depth: 5
    toc_float: yes
  word_document:
    toc: yes
    toc_depth: '5'
  pdf_document:
    latex_engine: lualatex
    number_sections: yes
    toc: yes
    toc_depth: 5
---


```{r setup, include=FALSE, cache = FALSE}
require("knitr")
opts_chunk$set(tidy.opts=list(width.cutoff=60),tidy=TRUE)
opts_chunk$set(engine = "R")
knitr::opts_chunk$set(echo = T)
knitr::opts_chunk$set(message = F)
knitr::opts_chunk$set(warning = F)
## setting wd in DELL
## opts_knit$set(root.dir = "~/Stats/CIDA_OMICs/CIDA_OMICS/7659Stats_Genetics/HW5/")
## setting working directory in asus 
## opts_knit$set(root.dir = "C:/Users/hithr/Documents/Stats/gitlab/Cario_RNASeq_Microbiom_Inte/DataRaw/") 
## setting working directory in ubuntu
opts_knit$set(root.dir = "~/Documents/gitlab/Cario_RNASeq_Microbiom_Inte/DataRaw/")
                                                 
## cache = F, if cache = T, will not revaluate code chunk everytime
## double or more space to insert a line break
```


```{r libs, warning=FALSE, echo=F}
######################
## Set up workspace
######################
rm(list = ls())
library(edgeR)
library(EDASeq)
library(DESeq2)
library(knitr)
library(tidyverse)
library(magrittr)
library(stats)
library(BiocParallel)
library(readxl)
library(openxlsx)
library(limma)
library(scater)
library(rgl)
library(pca3d)
library(EnhancedVolcano)
library(RColorBrewer)
library(grid)
library(cowplot)
require("ggrepel")
options(stringsAsFactors = F)
options(dplyr.width = Inf)
getwd()
## not in function
'%nin%' <- Negate('%in%')

# ######## clean memory ######################
# rm(list = ls())
# gc()
# is(dds)
# slotNames(dds)

```


```{r edger}
## data 
# import unnormalized counts table
cnts.raw <- read.delim("All_Sample_geneCounts_raw_counts.txt", header = TRUE, sep = "\t")

cnts <- cnts.raw %>% dplyr::select(-c(Symbol, Length ) ) %>% tibble::column_to_rownames("Gene_ID")
cnts <- as.matrix(cnts)
rna.pid <- colnames(cnts)
# now we have the common counts table 
## pheno
ctrl.id <-  colnames(cnts)[1:13]
hiv.id <- colnames(cnts)[14:32]
## from dim() we know there are 32 samples
pheno <- data.frame(pid = rna.pid, txt = as.factor(c(rep("Control", 13), 
                                                     rep("HIV", 19)
                                                     )) )
pheno$txt %<>% relevel("Control")

## add Symbol to the data  
cnts <- as.data.frame(cnts)
cnts$Symbol <- cnts.raw$Symbol
cnts$Length <- cnts.raw$Length

##filter the raw data and check dim
cnts_fsym <- cnts[rowSums(cnts[, 1:32])>=(5*ncol(cnts[, 1:32])), ]
## should end up around 15 - 20K genes 
ngenes <- nrow(cnts_fsym)
paste("The number of remaining genes: ", ngenes, sep = '')
# all integer in cnts_f
cnts_f <- as.matrix(cnts_fsym[, 1:32])

# using the function from EDASeq
set <- newSeqExpressionSet(as.matrix(cnts_f),phenoData = data.frame(condition=as.factor(pheno$txt), row.names=colnames(cnts_f)))

########### DESeq 2 ###################
# using the function from EDASeq
set <- newSeqExpressionSet(as.matrix(cnts_f),phenoData = data.frame(condition=as.factor(pheno$txt), row.names=colnames(cnts_f)))
## now using deseq2
dds <- DESeqDataSetFromMatrix(countData = counts(set), colData = pData(set),design = ~ condition)
dds <- estimateSizeFactors(dds)
## normalization factors
sizeFactors(dds)
cnts.deseq2 <- counts(dds, normalized=TRUE)

############### edger wrapper #########################
## edger_wrapper <- function()

######## edgeR TMM ###############
## edgeR object
group <- c(rep(1, 13), rep(2, 19))
y <- DGEList(counts= as.matrix(cnts_f), group=group)
## normalization 
y <- calcNormFactors(y, method = "TMM")
y$samples
cnts.edger <- edgeR::cpm(y)


## DE analysis of edger
y <- estimateDisp(y)
et <- exactTest(y, pair = 1:2)
dim(et)
### results
topTags(et, n = 10, adjust.method = "BH")
head(et$table)
et$comparison
## summary results
summary(decideTests(et, p.value = 0.05,
            lfc = 0))
paste(1706+ 2540, "genes with FDR 0.05")
summary(decideTests(et, p.value = 0.05,
            lfc = 1))
paste(157 + 680, "genes with FDR 0.05")
summary(decideTests(et, p.value = 0.05,
            lfc = 2))

## results summary 
res.edger <- et$table %>% dplyr::mutate(Symbol = cnts_fsym$Symbol,
                                 FDR = p.adjust(et$table$PValue, method = "BH") )

```


```{r plot_norm_counts}
## plot counts, with source code from plotCounts()
group.plot <- factor(c(rep(1,13), rep(2,19)), labels = c("Health Control", "HIV Infected"))


## use the # No. of genes in the cnts.edger matrix
sim_plotcounts <- function(genenumber){
  plot(as.integer(group.plot)+ runif(ncol(cnts.edger), -0.1, 0.1), cnts.edger[genenumber, ], 
    xlim = c(0.5, length(levels(group.plot)) + 0.8), xaxt="n" ,
    xlab = "Groups", ylab = "Normalized Counts", main = res.edger[genenumber, ]$Symbol)
axis(1, at = seq_along(levels(group.plot)), levels(group.plot))
## pos below
text(length(levels(group.plot)) + 0.5, stats::quantile( cnts.edger[genenumber, ], 0.95), 
     paste("log2 Fold Change:", round(res.edger[genenumber, ]$logFC,2),
             "","Adjusted p-value:",  
           format(res.edger[genenumber, ]$FDR,digits=2,scientific=TRUE ),"",
           "Gene_ID: ", rownames(res.edger)[genenumber],
                                           sep = "\n"), pos = 1)
}
## plot top 10
for(i in order(res.edger$PValue)[1:10]){
  sim_plotcounts(i)
}


```


## Genes lists and their DE results
Clinical data, making sure the samples are in the same order as the samples' order in the genematrix.  

Significant altered genes within gene lists, 130 for IFN-beta and 117 for core ISGs. 

```{r clinical}
# clinical data, the samples are in the same order as the samples order in the genematrix
clinical_order <- read.csv("~/Documents/gitlab/Cario_RNASeq_Microbiom_Inte/DataProcessed/interferome/clinical_order.csv" )
clinical_names <- c("Blood CD4 T Cell Counts (cells/ul)", "Plasma Viral Load", "Tissue HIV RNA (per CD4 T cell)",
                    "Tissue CD4 T Cell Counts (number/g)", "IL-6 (pg/ml)", "CRP (ug/ml)", "iFABP (pg/ml)",
                    "sCD27 (U/ml)", "CD14 (ng/ml)", "LPS (pg/ml)", "LTA (OD)", 
                    base::paste("IFN", '\u03b1', sep = "" ),  base::paste("IFN", '\u03b2', sep = "" ))
n_clinical <- length(clinical_names)
clinical_names 
n_clinical
clinical_sum <- matrix(NA, 13,8)
# genelists
isgs <- as.data.frame(read.delim("coreISG"))
genesbeta <- as.data.frame(read.delim("genesbeta"))

# all DE results for genes lists ##
isgs_DE <- read.xlsx( "~/Documents/gitlab/Cario_RNASeq_Microbiom_Inte/DataProcessed/interferome/ISGs_DEanalysis_results.xlsx")
genesbeta_DE <- read.xlsx("~/Documents/gitlab/Cario_RNASeq_Microbiom_Inte/DataProcessed/interferome/IFNbeta_DEanalysis_results.xlsx")

############### cnts for genes lists fdr 0.05 ##################
setwd("~/Documents/gitlab/Cario_RNASeq_Microbiom_Inte/DataProcessed/")
cnts.genesbeta.edger.05 <- read.csv("corr/cnts.genesbeta.edger.05.csv")
cnts.isgs.edger.05 <- read.csv("corr/cnts.isgs.edger.05.csv")

## DE results
genesbeta.edger.05 <- read.csv("corr/genesbeta.edger.05.csv")
isgs.edger.05 <- read.csv("corr/isgs.edger.05.csv")


```

```{r de_scatter}


de_scatter <- function(de, label_fdr, label_fc, name){
  ggplot(de, aes(logFC, -log10(FDR) )) +
  geom_point(colour = "gray65") + 
  geom_point(data = subset(de, FDR < 5e-2 & logFC > 0 ),  aes(logFC, -log10(FDR) ), colour="black") +
  geom_point(data = subset(de, FDR < 5e-2 & logFC < 0 ),  aes(logFC, -log10(FDR) ), colour="gray35") +
  labs(x = expression(paste(log[2]," Fold Change" )),
        y =  expression(paste("-",log[10]," FDR" )),
       caption =  paste("Differential Expression Results of ", name) ) +
  geom_text_repel(data=subset(de, FDR < label_fdr & logFC > label_fc ),
             aes(label=Symbol), size = 3,
             box.padding = unit(0.35, "lines"),
              point.padding = unit(0.3, "lines")) +
  geom_hline(yintercept=-log10(0.05), linetype="dashed", color = "black") +
  geom_vline(xintercept=0, linetype="dashed", color = "black") +
  annotate(geom="text", x = 3, 
               y=-log10(0.01), 
            label="FDR < 0.05",
               color="black") +
  xlim(-1.5, 3.5) +
  theme_bw()
  
}


de_scatter(isgs_DE, 1e-5, 2, "Core ISGs")
de_scatter(genesbeta_DE , 1e-5, 2, "IFN-Beta Genes")

p1 <- de_scatter(isgs_DE, 1e-5, 2, "Core ISGs")
p2 <- de_scatter(genesbeta_DE , 1e-5, 2, "IFN-Beta Genes")
 grid.newpage()
 grid.draw(rbind(ggplotGrob(p1), ggplotGrob(p2), size = "last"))
  
```

## PCA and plots 

For all filered TMM normalized counts, significant altered genes in Core ISGs and IFN-beta genes. 


```{r pca_plots}
## check samples order
clinical_order <- read.csv("~/Documents/gitlab/Cario_RNASeq_Microbiom_Inte/DataProcessed/interferome/clinical_order.csv" )

## do pca and get some plots for the data (matrix)
# The data is subjects as columns and features at rows
symbol <- cnts_fsym$Symbol

## pheno
rna.pid <- colnames(cnts)[1:32]
pheno <- data.frame(pid = rna.pid, txt = as.factor(c(rep("Control", 13), 
                                                     rep("HIV", 19)
                                                     )) )
pheno$txt %<>% relevel("Control")

prcomp_wrapper <- function(data, gene_symbol, pheno, name){
  ####### default groups ##############
  group_labels = c("Healthy Controls",  "HIV-Infected")
  ###### PCA by prcomp ##########
  data.pca = prcomp(t(data), center = TRUE, scale. = TRUE, retx = TRUE)
  data.pca.symbol = data.pca$rotation
  row.names(data.pca.symbol) = gene_symbol
  data.pca.symbol = as.matrix(data.pca.symbol)
  ########## pc names ##############
  pc_sample = data.frame(data.pca$x)
  pc = colnames(data.pca$x)
  ######### variance of pcs############
  data.sd = ( data.pca$sdev/sum(data.pca$sdev) )* 100
  # data for bar plot
  pc.sd = data.frame(PCs = pc,
                      Variance = data.sd) %>% top_n(n = 9, wt = Variance)
  
  p.var = ggplot(data=pc.sd, aes(x=PCs, y=Variance)) +
   geom_bar(stat="identity", fill = "Black") +
    labs(y = "% Variance", 
         caption = paste("Variance of PCs for ", name ) ) +
   theme_bw()
   print(p.var)
   ########### pc1 vs pc2 #############
   pca12 = ggplot(data = pc_sample, aes(x = PC1, y = PC2, color = pheno$txt, shape = pheno$txt)) +
        geom_point(size = 2) +
        xlab(paste("PC1 (", round(pc.sd$Variance[1],2) ,"%", ")", 
                   sep = "") ) +
        ylab(paste("PC2 (", round(pc.sd$Variance[2],2) ,"%", ")", 
             sep = "") ) +
       scale_color_manual(values=c("gray70", "Black"), 
                    name="Groups",
                    labels = group_labels
                    ) +
      scale_shape_manual( values = c(16, 17), 
                    name="Groups",
                    labels = group_labels
                    ) +
       stat_ellipse(type = "t", linetype = "solid") +
       labs(caption = paste("PC1 vs PC2 for ", name ) ) +
       theme_bw()
       print(pca12)
       
      pca13 = ggplot(data = pc_sample, aes(x = PC1, y = PC3, color = pheno$txt, shape = pheno$txt)) +
        geom_point(size = 2) +
        xlab(paste("PC1 (", round(pc.sd$Variance[1],2) ,"%", ")", 
                   sep = "") ) +
        ylab(paste("PC3 (", round(pc.sd$Variance[3],2) ,"%", ")", 
             sep = "") ) +
       scale_color_manual(values=c("gray70", "Black"), 
                    name="Groups",
                    labels = group_labels
                    ) +
      scale_shape_manual( values = c(16, 17), 
                    name="Groups",
                    labels = group_labels
                    ) +
       stat_ellipse(type = "t", linetype = "solid") +
       labs(caption = paste("PC1 vs PC3 for ", name ) ) +
       theme_bw()
       print(pca13)
       
        pca23 = ggplot(data = pc_sample, aes(x = PC2, y = PC3, color = pheno$txt, shape = pheno$txt)) +
        geom_point(size = 2) +
        xlab(paste("PC2 (", round(pc.sd$Variance[2],2) ,"%", ")", 
                   sep = "") ) +
        ylab(paste("PC3 (", round(pc.sd$Variance[3],2) ,"%", ")", 
             sep = "") ) +
       scale_color_manual(values=c("gray70", "Black"), 
                    name="Groups",
                    labels = group_labels
                    ) +
      scale_shape_manual( values = c(16, 17), 
                    name="Groups",
                    labels = group_labels
                    ) +
       stat_ellipse(type = "t", linetype = "solid") +
       labs(caption = paste("PC2 vs PC3 for ", name ) ) +
       theme_bw()
       print(pca23)
    ############### pc1 vs il-6, ifn-beta ###############
    pc1_clin_1 = ggplot(mapping = aes(x = clinical_order$IL.6 , y = pc_sample$PC1) ) +
         labs(x = "IL-6 (pg/ml)",
              y = paste("PC1 (", round(pc.sd$Variance[1],2) ,"%", " Variance", ")", sep = "") , 
         caption = paste("IL-6 (pg/ml) vs PC1 for ", name ) )  +
         geom_point(aes(shape = pheno$txt, colour = pheno$txt), size = 2.5) +
        geom_smooth(method = "loess", se = FALSE, colour = "black") +
        geom_smooth(method = "lm", se = FALSE, colour = "black") +
        scale_color_manual(values=c("gray70", "Black"), 
                    name="Groups",
                    labels = group_labels
                    ) +
       scale_shape_manual( values = c(16, 17), 
                    name="Groups",
                    labels = group_labels
                    ) +
        theme_bw()
    print(pc1_clin_1)
    
   pc1_clin_2 = ggplot(mapping = aes(x = clinical_order$IFNb , y = pc_sample$PC1) ) +
         labs(x = "IFN-Beta",
              y = paste("PC1 (", round(pc.sd$Variance[1],2) ,"%", " Variance",")", sep = "") , 
         caption = paste("IFN-Beta vs PC1 for ", name ) )  +
         geom_point(aes(shape = pheno$txt, colour = pheno$txt), size = 2.5) +
        geom_smooth(method = "loess", se = FALSE, colour = "black") +
        geom_smooth(method = "lm", se = FALSE, colour = "black") +
        scale_color_manual(values=c("gray70", "Black"), 
                    name="Groups",
                    labels = group_labels
                    ) +
       scale_shape_manual( values = c(16, 17), 
                    name="Groups",
                    labels = group_labels
                    ) +
        theme_bw()
    print(pc1_clin_2)
}

prcomp_wrapper(cnts.edger, symbol, pheno, "TMM Normalized Counts")

##### altered genes lists ##############
cnts.isgs.edger.sig <- cnts.isgs.edger.05 %>% column_to_rownames("X") %>% as.matrix
symbol <- isgs.edger.05$Symbol.x
prcomp_wrapper(cnts.isgs.edger.sig, symbol, pheno, "Significantly Altered Core ISGs")


cnts.genesbeta.edger.sig <- cnts.genesbeta.edger.05 %>% column_to_rownames("X") %>% as.matrix
symbol <- genesbeta.edger.05$Symbol.x
prcomp_wrapper(cnts.genesbeta.edger.sig, symbol, pheno, "Significantly Altered IFN-Beta Genes")




```

The ellipse was drawn based on the multi-variate t distribution. 

### PC1 and Clinical Variables

```{r pc1_clin, echo = F}
pc_x <- function(data, gene_symbol) {
  ####### default groups ##############
  group_labels = c("Healthy Controls",  "HIV-Infected")
  ###### PCA by prcomp ##########
  data.pca = prcomp(t(data), center = TRUE, scale. = TRUE, retx = TRUE)
  data.pca.symbol = data.pca$rotation
  row.names(data.pca.symbol) = gene_symbol
  data.pca.symbol = as.matrix(data.pca.symbol)
  ########## pc names ##############
  pc_sample = data.frame(data.pca$x)
  return(pc_sample)
}

symbol <- isgs.edger.05$Symbol.x
isgs.pca.x <- pc_x (cnts.isgs.edger.sig, symbol)
symbol <- genesbeta.edger.05$Symbol.x
genesbeta.pca.x <- pc_x (cnts.genesbeta.edger.sig, symbol)

pc1_sum <- read.csv("~/Documents/gitlab/Cario_RNASeq_Microbiom_Inte/DataProcessed/corr/pc1_sum.csv")
pc1_sum <- as.data.frame(pc1_sum)
pc1_sum$p.value <- as.numeric(pc1_sum$p.value)
pc1_sum$Gene.Lists <- c( rep("PC1 of Significantly Altered Core ISGs" ,14)  ,
                          rep("PC1 of Significantly Altered IFN-Beta Genes" ,14))

ggplot(data = pc1_sum, aes(x=Clinical.Parameters, y= Gene.Lists, fill= -log10(p.value)  ) ) + 
  geom_tile(size = 2) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 50, vjust = 1, 
                                   size = 10, hjust = 1)) +
  coord_fixed() +
  xlab("") +
  ylab("") +
  scale_fill_gradient2(low = "Black", high = "red", mid = "white", 
                       midpoint = -log10(0.05), limit = c(0,  max(-log10(pc1_sum$p.value)) ), 
                       space = "Lab", 
                     name= expression(paste("-",log[10]," p value" )),
                     breaks = round(c(0, -log10(0.05), -log10(0.01), -log10(0.001)),2) ) +
  scale_color_discrete(name = "") 

kable(pc1_sum[, 1:3])

```



```{r test, echo=F, eval=F}

data = cnts.edger

data.pca = prcomp(t(data), center = TRUE, scale. = TRUE, retx = TRUE)
  data.pca.symbol = data.pca$rotation
  row.names(data.pca.symbol) = symbol
  data.pca.symbol = as.matrix(data.pca.symbol)
  ########## pc names ##############
  pc_sample = data.frame(data.pca$x)
  pc = colnames(data.pca$x)
  ######### variance of pcs############
  data.sd = ( data.pca$sdev/sum(data.pca$sdev) )* 100
  # data for bar plot
  pc.sd = data.frame(PCs = pc,
                      Variance = data.sd) %>% top_n(n = 9, wt = Variance)
  
  p.var = ggplot(data=pc.sd, aes(x=PCs, y=Variance)) +
   geom_bar(stat="identity", fill = "Black") +
    labs(y = "% Variance", 
         caption = paste("Variance of PCs for " ) ) +
   theme_bw()
   p.var

sum(row.names(pc_sample) == pheno$pid)


pca12 = ggplot(data = pc_sample, aes(x = PC1, y = PC2, color = pheno$txt, shape = pheno$txt)) +
        geom_point(size = 2) +
        xlab(paste("PC1 (", round(pc.sd$Variance[1],2) ,"%", ")", 
                   sep = "") ) +
        ylab(paste("PC2 (", round(pc.sd$Variance[2],2) ,"%", ")", 
             sep = "") ) +
       scale_color_manual(values=c("gray70", "Black"), 
                    name="Groups",
                    labels=c("Healthy Controls",  "HIV-Infected")
                    ) +
      scale_shape_manual( values = c(16, 17), 
                    name="Groups",
                    labels=c("Healthy Controls",  "HIV-Infected")
                    ) +
       stat_ellipse(type = "t", linetype = "solid") +
       labs(caption = paste("Variance of PCs for " ) ) +
       theme_bw()
pca12
  

######## check sample #############
clinical_order$pid == pheno$pid

## check samples ######
colnames(cnts.isgs.edger.05)[-1] == colnames(cnts.edger)
colnames(cnts.genesbeta.edger.05)[-1] == colnames(cnts.edger)

sum(isgs.edger.05$Gene_ID == cnts.isgs.edger.05$X)
sum(genesbeta.edger.05$Gene_ID == cnts.genesbeta.edger.05$X)

####### get pc1 #####
# pc_sample

```